---
title: "Trawl vs. eDNA Metabarcoding: Cephalopod Survey Comparison"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

# Overview

This report harmonizes paired trawl survey and eDNA metabarcoding datasets collected at the same stations, then tests:

1. **Rank correlation** (Spearman's ρ) between trawl catch and eDNA read abundance
2. **Relative abundance correlation** across all paired station × species observations
3. **Differences among eDNA sample types** (e.g. filter type, replicate type) in their correlation with trawl data

---

# 1. Setup & Data Loading

```{r setup}
library(tidyverse)
library(data.table)
library(vegan)
library(ggplot2)
library(ggrepel)
library(ggtext)
library(ggvenn)
library(patchwork)
library(knitr)
library(kableExtra)
library(broom)

# ── USER: Update these file paths ────────────────────────────────────────────
trawl_file       <- "inputs/trawl_taxa_subset.csv"           # Long-format trawl CSV
edna_spp_files   <- c(                          # One entry per sequencing run —
  run1 = "inputs/SPECIES_counts_Trawl_Ceph18S_Run1.csv",        #   named with a short run label
  run2 = "inputs/SPECIES_counts_Trawl_Ceph18S_Run2.csv"        #   named with a short run label
)
edna_meta_file   <- "inputs/Metabarcoding_Metadata.csv"     # Sample metadata — must cover ALL runs
lookup_file      <- "inputs/species_types.csv"              # Common-name ↔ binomial lookup table

# ── USER: Update these column names to match your files ──────────────────────
# Trawl columns
trawl_station_col  <- "TowID"          # Station/site identifier
trawl_species_col  <- "Common_Name"    # Common name column in trawl
trawl_count_col    <- "Number_Caught"  # Individual counts  — set to NA if column absent
trawl_weight_col   <- "Weight_kg"      # Biomass column     — set to NA if column absent

# eDNA species table columns (same structure expected in every run file)
edna_species_col   <- "Species"        # Binomial species name column
edna_tax_col       <- "full_tax"       # Full taxonomy string column

# eDNA metadata columns
meta_sample_col    <- "ReadFile"       # Unique sample ID matching column names in eDNA tables
                                       #   NOTE: must match the sample column headers in the species table
meta_station_col   <- "TrawlID"        # Station identifier (matches trawl station IDs)
meta_type_col      <- "Sample_type"    # Column distinguishing sample types
meta_run_col       <- "RunID"          # Sequencing run identifier column in metadata
meta_primer_col    <- "Primer"         # Primer/amplicon column in metadata (set to NA if absent)

# ── USER: Primer filter ───────────────────────────────────────────────────────
# Restrict to samples sequenced with specific primer(s). Set to NULL to keep all.
# Examples:
#   Ceph18S  → valid_primers <- c("Ceph18S")
#   MiFish   → valid_primers <- c("MiFish12S")
#   LerayCOI → valid_primers <- c("LerayCOI")
valid_primers <- c("Ceph18S")        # ← change to your primer name(s) or leave NULL
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Sample types to keep ───────────────────────────────────────────────
# List the valid environmental sample types (all others — controls, mocks, NTCs — are excluded).
# Examples:
#   Ceph18S / MiFish → valid_sample_types <- c("Bottom_metaprobe", "Top_metaprobe", "Slush")
#   LerayCOI         → valid_sample_types <- c("eDNA_filter", "water_sample")   # adjust as needed
valid_sample_types <- c("Bottom_metaprobe", "Top_metaprobe", "Slush")
# ─────────────────────────────────────────────────────────────────────────────

# Station ID value used for controls/blanks that should be excluded
invalid_station_values <- c("NA", "", "na")

# Lookup table columns (species_types.csv)
lookup_common_col  <- "Common_Name"    # Common name  → matches trawl species
lookup_binom_col   <- "Species"        # Binomial name → matches eDNA species
lookup_group_col   <- "Fish_Invert"    # Taxonomic group (Fish / Invertebrate)
lookup_type_col    <- "Body_Type"      # Body type (Slimy / Crunchy)

# ── USER: Stations to exclude ────────────────────────────────────────────────
# List any trawl station IDs to remove from ALL analyses (trawl + eDNA).
# Useful when eDNA samples failed at a station and you want to drop it entirely.
# Set to NULL to keep all stations.
exclude_stations <- NULL     # e.g. c("FL23072", "FL23101") or NULL
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Species pooling map ─────────────────────────────────────────────────
# Some loci cannot distinguish species within a genus. Use this to pool them.
# Named character vector: names are regex patterns (anchored to start), values
# are the pooled label. The patterns are matched case-insensitively.
# Set to NULL to skip species pooling.
#
# Examples:
#   MiFish12S  → pool_species_map <- c("^Alosa "        = "Alosa spp.",
#                                       "^Myoxocephalus " = "Myoxocephalus spp.")
#   Ceph18S    → pool_species_map <- NULL  (species are distinguishable)
#   LerayCOI   → pool_species_map <- NULL  (or add any genus-level ambiguities)
pool_species_map <- NULL     # ← set to named character vector or NULL
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Taxon group filter ──────────────────────────────────────────────────
# Restrict all analyses to one taxonomic group. Options:
#   "Fish"          — keep only rows where taxon_group == "Fish"
#   "Invertebrate"  — keep only rows where taxon_group == "Invertebrate"
#   "Cephalopod"    — keep only the species listed in cephalopod_species below
#   NULL            — no filter; use all species
taxon_filter <- "Cephalopod"   # ← set to "Fish", "Invertebrate", "Cephalopod", or NULL

# Species included when taxon_filter == "Cephalopod"
cephalopod_species <- c("Illex illecebrosus", "Doryteuthis pealeii", "Bathypolypus arcticus")
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Sample type pooling strategy for Spearman correlations ─────────────
# How to pool eDNA reads across sample types for the "Pooled" Spearman analysis:
#   "max_reads"  — per station, keep only the sample_type with the highest total
#                  read count. Preserves the strongest signal; recommended.
#   "mean"       — average edna_reads across sample types per station × species.
#                  Conservative; reduces influence of any single sample type.
pool_strategy <- "mean"   # ← "max_reads" or "mean"
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Read and sample quality filters ────────────────────────────────────
min_reads_per_taxon  <- 20      # Remove taxa with fewer than this many raw reads (per run)
min_reads_per_sample <- 1000    # Drop samples with fewer than this many total reads
# ─────────────────────────────────────────────────────────────────────────────
```

```{r load-data}
trawl_raw  <- read_csv(trawl_file,    show_col_types = FALSE)
meta_raw   <- read_csv(edna_meta_file, show_col_types = FALSE)
lookup_raw <- read_csv(lookup_file,    show_col_types = FALSE)

# Load each sequencing run file and tag with its run label
edna_spp_raw_list <- imap(edna_spp_files, function(path, run_label) {
  df <- read_csv(path, show_col_types = FALSE)
  df$.run <- run_label
  df
})

cat("Trawl rows:", nrow(trawl_raw), "\n")
cat("eDNA metadata rows:", nrow(meta_raw), "\n")
iwalk(edna_spp_raw_list, ~ cat("Run", .y, "—", nrow(.x), "species ×",
                                ncol(.x) - 1, "columns\n"))
cat("Lookup table rows:", nrow(lookup_raw), "\n")
```

---

# 2. Data Harmonization

## 2a. Trawl Data — Clean & Summarize

```{r trawl-clean}
trawl <- trawl_raw %>%
  rename(
    station     = all_of(trawl_station_col),
    common_name = all_of(trawl_species_col)
  ) %>%
  mutate(common_name = str_squish(common_name))

# Retain both count and biomass columns where available
has_count  <- !is.na(trawl_count_col)  && trawl_count_col  %in% names(trawl)
has_weight <- !is.na(trawl_weight_col) && trawl_weight_col %in% names(trawl)

if (!has_count && !has_weight)
  stop("Neither count nor weight column found in trawl data.")

if (has_count)  trawl <- trawl %>% rename(trawl_count  = all_of(trawl_count_col))
if (has_weight) trawl <- trawl %>% rename(trawl_weight = all_of(trawl_weight_col))

# Aggregate in case of duplicates within station × species
trawl_agg <- trawl %>%
  group_by(station, common_name) %>%
  summarise(
    across(any_of(c("trawl_count", "trawl_weight")),
           ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

cat("Trawl stations:", n_distinct(trawl_agg$station), "\n")
cat("Trawl species (common names):", n_distinct(trawl_agg$common_name), "\n")
cat("Count column present:", has_count, "| Weight column present:", has_weight, "\n")
```

## 2b. eDNA Data — Load & Combine Sequencing Runs

Each run file has the same structure (species as rows, sample IDs as columns, plus
`Species` and `full_tax` annotation columns). Runs are pivoted to long format individually
then row-bound, retaining a `seq_run` tag on every row.

```{r edna-load}
annot_cols <- c(edna_species_col, edna_tax_col)

edna_long_runs <- map(edna_spp_raw_list, function(df) {
  run_label  <- unique(df$.run)
  sample_cols <- setdiff(names(df), c(annot_cols, ".run"))

  df_long <- df %>%
    filter(
      !is.na(.data[[edna_species_col]]),
      !tolower(str_squish(.data[[edna_species_col]])) %in% c("na", "")
    ) %>%
    mutate(species = str_to_sentence(str_squish(.data[[edna_species_col]]))) %>%
    select(species, full_tax = all_of(edna_tax_col), all_of(sample_cols)) %>%
    pivot_longer(
      cols      = all_of(sample_cols),
      names_to  = "sample_id",
      values_to = "reads"
    ) %>%
    filter(reads > 0) %>%
    mutate(seq_run = run_label)

  # ── Filter 1: remove taxa with fewer than min_reads_per_taxon reads ──────────
  # (raw reads, before TSS normalization — applied per-taxon across all samples
  #  within this run, to remove very low-confidence assignments)
  taxa_totals <- df_long %>%
    group_by(species) %>%
    summarise(total_reads_taxon = sum(reads, na.rm = TRUE), .groups = "drop")

  n_before_taxa <- n_distinct(df_long$species)
  df_long <- df_long %>%
    left_join(taxa_totals, by = "species") %>%
    filter(total_reads_taxon >= min_reads_per_taxon) %>%
    select(-total_reads_taxon)
  n_after_taxa <- n_distinct(df_long$species)
  cat("Run", run_label, "— taxa removed (< ", min_reads_per_taxon, " reads):",
      n_before_taxa - n_after_taxa, "| taxa retained:", n_after_taxa, "\n")

  # ── Filter 2: drop samples with fewer than min_reads_per_sample total reads ──
  sample_totals_run <- df_long %>%
    group_by(sample_id) %>%
    summarise(total_reads_sample = sum(reads, na.rm = TRUE), .groups = "drop")

  n_before_samp <- n_distinct(df_long$sample_id)
  df_long <- df_long %>%
    left_join(sample_totals_run, by = "sample_id") %>%
    filter(total_reads_sample >= min_reads_per_sample) %>%
    select(-total_reads_sample)
  n_after_samp <- n_distinct(df_long$sample_id)
  cat("Run", run_label, "— samples removed (< ", min_reads_per_sample, " reads):",
      n_before_samp - n_after_samp, "| samples retained:", n_after_samp, "\n")

  df_long
})

edna_long_all <- bind_rows(edna_long_runs)

cat("Total eDNA samples across all runs:", n_distinct(edna_long_all$sample_id), "\n")
cat("eDNA species (assigned):", n_distinct(edna_long_all$species), "\n")
cat("Non-zero observations:", nrow(edna_long_all), "\n")
iwalk(edna_long_runs, ~ cat("  Run", .y, "—",
                             n_distinct(.x$sample_id), "samples,",
                             n_distinct(.x$species), "species\n"))
```

## 2c. Run Normalization & Duplicate Sample Resolution

**Why normalize?** Read depth varies between sequencing runs, so raw read counts are not
directly comparable across runs. We use Total Sum Scaling (TSS) — dividing each sample's
reads by its total reads — to convert to within-sample proportions before combining runs.
Downstream analyses use these normalized values; the heatmaps display them as proportions
rather than raw counts.

**Why deduplicate?** If the same station × sample_type combination appears in both runs,
we keep only one representative sample to avoid pseudo-replication in the correlation
analyses.

```{r run-normalize-dedup}
# ── Step 1: attach metadata to get station + sample_type per sample ───────────
meta_clean_pre <- meta_raw %>%
  rename(
    sample_id   = all_of(meta_sample_col),   # ReadFile — matches species table columns
    station     = all_of(meta_station_col),   # TrawlID
    sample_type = all_of(meta_type_col)       # Sample_type
  ) %>%
  # Filter to the target primer(s) if specified
  { if (!is.null(valid_primers) && meta_primer_col %in% names(.))
      filter(., .data[[meta_primer_col]] %in% valid_primers)
    else . } %>%
  # Filter out controls, mocks, NTCs and rows with invalid station IDs
  filter(
    sample_type %in% valid_sample_types,
    !is.na(station),
    !tolower(str_squish(station)) %in% tolower(invalid_station_values)
  )

# Report primer filtering
if (!is.null(valid_primers) && meta_primer_col %in% names(meta_raw)) {
  all_primers <- sort(unique(meta_raw[[meta_primer_col]]))
  cat("Primers in metadata:", paste(all_primers, collapse = ", "), "\n")
  cat("Primers kept:", paste(valid_primers, collapse = ", "), "\n")
}
cat("Metadata rows after filtering controls + primer:", nrow(meta_clean_pre), "\n")
cat("Sample types retained:", paste(sort(unique(meta_clean_pre$sample_type)), collapse = ", "), "\n")
cat("Stations:", paste(sort(unique(meta_clean_pre$station)), collapse = ", "), "\n")

# Attach seq_run from metadata if column exists; otherwise use the run tag from the file
if (meta_run_col %in% names(meta_raw)) {
  meta_clean_pre <- meta_clean_pre %>%
    rename(seq_run_meta = all_of(meta_run_col))
} else {
  meta_clean_pre <- meta_clean_pre %>% mutate(seq_run_meta = NA_character_)
}

edna_with_meta <- edna_long_all %>%
  left_join(meta_clean_pre %>%
              select(sample_id, station, sample_type, seq_run_meta),
            by = "sample_id") %>%
  mutate(seq_run = coalesce(seq_run_meta, seq_run)) %>%
  select(-seq_run_meta) %>%
  # Drop any samples that didn't match valid metadata (controls etc.)
  filter(!is.na(station), !is.na(sample_type))

cat("\neDNA observations after joining metadata:", nrow(edna_with_meta), "\n")
cat("eDNA samples matched to valid trawl stations:",
    n_distinct(edna_with_meta$sample_id), "\n")

# ── Step 2: TSS normalization within each sample ──────────────────────────────
sample_totals <- edna_with_meta %>%
  group_by(sample_id, seq_run) %>%
  summarise(total_reads = sum(reads, na.rm = TRUE), .groups = "drop")

edna_norm <- edna_with_meta %>%
  left_join(sample_totals, by = c("sample_id", "seq_run")) %>%
  mutate(reads_tss = reads / total_reads)   # proportion within sample

cat("TSS normalization check — NA total_reads:", sum(is.na(edna_norm$total_reads)), "\n")

# ── Step 3: identify duplicate station × sample_type combos across runs ──────
sample_index <- edna_norm %>%
  distinct(sample_id, station, sample_type, seq_run) %>%
  filter(!is.na(station), !is.na(sample_type))

# Count species per sample (used for representative_sample strategy)
species_per_sample <- edna_norm %>%
  group_by(sample_id) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

# Find station × sample_type groups that span more than one sequencing run
run_per_group <- sample_index %>%
  group_by(station, sample_type) %>%
  summarise(runs_present = list(sort(unique(seq_run))),
            n_runs = n_distinct(seq_run), .groups = "drop")

dup_groups <- run_per_group %>% filter(n_runs > 1)

if (nrow(dup_groups) == 0) {
  cat("✓ No station × sample_type combinations span multiple runs.\n")
  edna_long_representative <- edna_norm
  edna_long_average        <- edna_norm
} else {
  cat("⚠", nrow(dup_groups),
      "station × sample_type combinations have samples in >1 run:\n")
  print(dup_groups %>% select(station, sample_type, n_runs))

  # ── Strategy A: representative_sample ────────────────────────────────────────
  # For each duplicated station × sample_type, keep the SAMPLE with the most
  # species (highest diversity). Within a run's replicates, pick the best one
  # (most species) as the representative; then use just that one sample.
  # Non-duplicated groups: keep all samples (replicates will be averaged later).
  run_depth <- sample_index %>%
    left_join(sample_totals,     by = c("sample_id", "seq_run")) %>%
    left_join(species_per_sample, by = "sample_id") %>%
    group_by(station, sample_type, seq_run) %>%
    slice_max(n_species, n = 1, with_ties = FALSE) %>%
    summarise(
      best_sample_id    = sample_id,
      group_n_species   = n_species,
      group_total_reads = total_reads,
      .groups = "drop"
    )

  preferred_run_per_group_repr <- run_depth %>%
    group_by(station, sample_type) %>%
    slice_max(group_n_species, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(station, sample_type,
           preferred_run   = seq_run,
           preferred_sample = best_sample_id)

  cat("\nrepresentative_sample: keeping the sample with most species per dup group.\n")

  edna_long_representative <- edna_norm %>%
    left_join(preferred_run_per_group_repr,
              by = c("station", "sample_type")) %>%
    filter(
      is.na(preferred_run) |
      (seq_run == preferred_run & sample_id == preferred_sample)
    ) %>%
    select(-preferred_run, -preferred_sample)

  cat("Samples retained (representative_sample):",
      n_distinct(edna_long_representative$sample_id), "/",
      n_distinct(edna_norm$sample_id), "\n")

  # ── Strategy B: average ───────────────────────────────────────────────────────
  # TSS-normalize each sample (already done in edna_norm), then average reads_tss
  # across ALL samples for the same station × sample_type × species — even across
  # runs. Averaging happens in edna_agg (section 2e).
  edna_long_average <- edna_norm
  cat("average: all samples retained (", n_distinct(edna_long_average$sample_id),
      "); duplicates averaged in edna_agg\n", sep = "")
}

# Summary of run composition
for (strategy_label in c("representative_sample", "average")) {
  elong <- if (strategy_label == "representative_sample") edna_long_representative else edna_long_average
  run_summary <- elong %>%
    distinct(sample_id, station, sample_type, seq_run) %>%
    filter(!is.na(station)) %>%
    count(seq_run, name = "n_samples")
  cat("\nSamples per sequencing run [", strategy_label, "]:\n", sep = "")
  print(run_summary)
}

# ── Active edna_long alias (updated after station exclusion below) ────────────
edna_long <- edna_long_representative
```

## 2d. Species Name Harmonization via Lookup Table

The trawl uses common names; the eDNA table uses binomial names. The lookup table
(`species_types.csv`) bridges them. This chunk adds the binomial key to trawl records
and flags any common names not found in the lookup.

```{r lookup-join}
# Clean lookup table
lookup <- lookup_raw %>%
  rename(
    common_name  = all_of(lookup_common_col),
    species      = all_of(lookup_binom_col),
    taxon_group  = all_of(lookup_group_col),
    body_type    = all_of(lookup_type_col)
  ) %>%
  mutate(
    common_name = str_squish(common_name),
    species     = str_to_sentence(str_squish(species))
  ) %>%
  select(common_name, species, taxon_group, body_type)

# Join binomial name onto trawl aggregated data
trawl_agg <- trawl_agg %>%
  left_join(lookup, by = "common_name")

# Report any trawl common names missing from the lookup
unmatched_trawl <- trawl_agg %>%
  filter(is.na(species)) %>%
  distinct(common_name)

if (nrow(unmatched_trawl) > 0) {
  message("⚠ These trawl common names were NOT found in the lookup table and will be excluded from correlations:")
  print(unmatched_trawl)
} else {
  cat("✓ All trawl common names matched in lookup table.\n")
}

# Report eDNA binomials not in lookup (across all runs)
all_edna_species <- map(edna_spp_raw_list, ~ {
  .x %>%
    pull(.data[[edna_species_col]]) %>%
    keep(~ !is.na(.x) && !tolower(str_squish(.x)) %in% c("na", "")) %>%
    str_to_sentence() %>% str_squish()
}) %>% unlist() %>% unique()

unmatched_edna <- setdiff(all_edna_species, lookup$species)
cat("\neDNA species not in lookup table (", length(unmatched_edna), "):",
    paste(head(unmatched_edna, 10), collapse = ", "),
    if (length(unmatched_edna) > 10) "..." else "", "\n")

# Keep only trawl records with a matched binomial
trawl_agg <- trawl_agg %>% filter(!is.na(species))

# ── Taxon group filter ────────────────────────────────────────────────────────
if (!is.null(taxon_filter)) {
  if (taxon_filter == "Cephalopod") {
    ceph_clean <- str_to_sentence(str_squish(cephalopod_species))
    trawl_agg  <- trawl_agg %>%
      filter(str_to_sentence(str_squish(species)) %in% ceph_clean)
    cat("Taxon filter: Cephalopod —", nrow(trawl_agg), "trawl records retained\n")
    cat("Species kept:", paste(sort(unique(trawl_agg$species)), collapse = ", "), "\n")
  } else {
    trawl_agg <- trawl_agg %>%
      filter(str_to_sentence(taxon_group) == str_to_sentence(taxon_filter))
    cat("Taxon filter:", taxon_filter, "—", nrow(trawl_agg), "trawl records retained\n")
    cat("Species kept:", paste(sort(unique(trawl_agg$species)), collapse = ", "), "\n")
  }
} else {
  cat("No taxon filter applied — all", nrow(trawl_agg), "trawl records retained\n")
}

# Derive the group label used in plot titles
taxon_label <- if (is.null(taxon_filter)) "All species" else taxon_filter
```

## 2d-i. Species Pooling for Indistinguishable Taxa

Some loci cannot reliably distinguish species within a genus. If `pool_species_map` is set,
ambiguous species are collapsed to genus-level labels (e.g. *Alosa spp.*) in both trawl
and eDNA data before any further analyses.

```{r pool-species}
# ── Helper: pool species names using pool_species_map ─────────────────────────
# pool_species_map is a named character vector:
#   names  = regex patterns matched against species names (case-insensitive)
#   values = replacement pooled label
# Example: c("^Alosa " = "Alosa spp.", "^Myoxocephalus " = "Myoxocephalus spp.")
# If pool_species_map is NULL, pool_genus() is a no-op.
pool_genus <- function(sp_vec) {
  if (is.null(pool_species_map) || length(pool_species_map) == 0) return(sp_vec)
  result <- sp_vec
  for (pattern in names(pool_species_map)) {
    replacement <- pool_species_map[[pattern]]
    result <- if_else(str_detect(result, regex(pattern, ignore_case = TRUE)),
                      replacement, result)
  }
  result
}

if (!is.null(pool_species_map) && length(pool_species_map) > 0) {

  # ── Pool in trawl_agg ─────────────────────────────────────────────────────────
  trawl_agg_pre <- trawl_agg
  trawl_agg <- trawl_agg %>%
    mutate(species = pool_genus(species)) %>%
    group_by(station, species, common_name, taxon_group, body_type) %>%
    summarise(
      across(any_of(c("trawl_count", "trawl_weight")), ~ sum(.x, na.rm = TRUE)),
      .groups = "drop"
    )

  pooled_trawl <- setdiff(unique(trawl_agg_pre$species), unique(trawl_agg$species))
  cat("Trawl species pooled:", paste(pooled_trawl, collapse = ", "), "\n")
  cat("Trawl species after pooling:", n_distinct(trawl_agg$species), "\n")

  # ── Pool in both edna_long strategies ────────────────────────────────────────
  edna_long_pre_pool_n <- n_distinct(edna_long_representative$species)

  edna_long_representative <- edna_long_representative %>%
    mutate(species = pool_genus(species))

  edna_long_average <- edna_long_average %>%
    mutate(species = pool_genus(species))

  edna_long <- edna_long_representative

  cat("eDNA species after pooling:", n_distinct(edna_long_representative$species),
      "(was", edna_long_pre_pool_n, ")\n")

  # ── Extend lookup to include pooled names ─────────────────────────────────────
  # Derive pooled genus names from the map values
  pooled_names <- unique(unname(pool_species_map))

  lookup_spp_extra <- lookup %>%
    filter(pool_genus(species) %in% pooled_names, species != pool_genus(species)) %>%
    mutate(species = pool_genus(species)) %>%
    distinct(species, taxon_group, body_type) %>%
    mutate(common_name = species)

  lookup <- bind_rows(
    lookup %>% filter(!species %in% lookup_spp_extra$species),
    lookup_spp_extra
  )
  cat("Lookup entries after adding pooled rows:", nrow(lookup), "\n")

} else {
  cat("No species pooling applied (pool_species_map is NULL).\n")
}
```

## 2d-ii. Station Exclusion

```{r exclude-stations}
if (!is.null(exclude_stations) && length(exclude_stations) > 0) {
  n_trawl_before <- n_distinct(trawl_agg$station)
  n_repr_before  <- n_distinct(edna_long_representative$station)
  n_avg_before   <- n_distinct(edna_long_average$station)

  trawl_agg              <- trawl_agg              %>% filter(!station %in% exclude_stations)
  edna_long_representative <- edna_long_representative %>% filter(!station %in% exclude_stations)
  edna_long_average        <- edna_long_average        %>% filter(!station %in% exclude_stations)
  edna_long                <- edna_long_representative  # update alias

  cat("Excluded stations:", paste(exclude_stations, collapse = ", "), "\n")
  cat("Trawl stations removed:", n_trawl_before - n_distinct(trawl_agg$station),
      "| remaining:", n_distinct(trawl_agg$station), "\n")
  cat("eDNA (representative) stations after exclusion:", n_distinct(edna_long_representative$station), "\n")
  cat("eDNA (average) stations after exclusion:", n_distinct(edna_long_average$station), "\n")
} else {
  cat("No stations excluded.\n")
}
```

## 2e–8. Dual-Strategy Analysis

All analyses from eDNA aggregation through CSV export are run twice — once for each
normalization/dedup strategy:

- **representative_sample**: TSS-normalize each sample, then for any duplicate
  station × sample_type across runs, keep the single sample with the most species.
- **average**: TSS-normalize each sample, then average `reads_tss` across all samples
  (including duplicates) for the same station × sample_type × species.

Results are labelled with the strategy name throughout.

```{r dual-strategy-loop, results='asis'}
# ── Metadata warning (once, not per strategy) ─────────────────────────────────
meta_clean <- meta_clean_pre
missing_meta <- setdiff(unique(edna_long_all$sample_id), meta_clean$sample_id)
if (length(missing_meta) > 0)
  warning("These sample IDs are in the eDNA table but not in metadata: ",
          paste(missing_meta, collapse = ", "))

# ── Loop over both strategies ─────────────────────────────────────────────────
strategies <- list(
  representative_sample = edna_long_representative,
  average               = edna_long_average
)

for (strategy_name in names(strategies)) {
  edna_long_strat <- strategies[[strategy_name]]

  cat("\n\n---\n\n")
  cat("## Strategy:", strategy_name, "\n\n")

  # ── 2e. eDNA aggregation ────────────────────────────────────────────────────
  # For 'average': mean() collapses both intra-run replicates and cross-run dups.
  # For 'representative_sample': only one sample per dup group, so mean() = that sample.
  edna_agg <- edna_long_strat %>%
    filter(!is.na(station), !is.na(sample_type)) %>%
    group_by(station, sample_type, species) %>%
    summarise(edna_reads = mean(reads_tss, na.rm = TRUE), .groups = "drop")

  cat("eDNA sample types:", paste(sort(unique(edna_agg$sample_type)), collapse = ", "), "\n")
  cat("eDNA stations:", n_distinct(edna_agg$station), "\n")

  # ── 2f. Join Trawl & eDNA ───────────────────────────────────────────────────
  shared_stations <- intersect(unique(trawl_agg$station), unique(edna_agg$station))
  cat("Paired stations (trawl ∩ eDNA):", length(shared_stations), "\n")

  paired <- trawl_agg %>%
    filter(station %in% shared_stations) %>%
    full_join(
      edna_agg %>% filter(station %in% shared_stations),
      by = c("station", "species")
    ) %>%
    mutate(
      across(any_of(c("trawl_count", "trawl_weight")), ~ replace_na(.x, 0)),
      edna_reads = replace_na(edna_reads, 0)
    ) %>%
    left_join(lookup %>% select(species, taxon_group, body_type), by = "species")

  paired <- paired %>%
    group_by(station, sample_type) %>%
    mutate(edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)) %>%
    ungroup()

  if (has_count) {
    paired <- paired %>%
      group_by(station, sample_type) %>%
      mutate(trawl_count_rel = trawl_count / sum(trawl_count, na.rm = TRUE)) %>%
      ungroup()
  }
  if (has_weight) {
    paired <- paired %>%
      group_by(station, sample_type) %>%
      mutate(trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) %>%
      ungroup()
  }

  both_detected <- paired %>% filter(edna_reads > 0)
  if (has_count)  both_detected <- both_detected %>% filter(trawl_count  > 0)
  if (has_weight) both_detected <- both_detected %>% filter(trawl_weight > 0)
  cat("Species × station pairs detected by both methods:", nrow(both_detected), "\n")

  # ── Pooled eDNA ─────────────────────────────────────────────────────────────
  if (pool_strategy == "max_reads") {
    best_type <- edna_agg %>%
      group_by(station, sample_type) %>%
      summarise(total_reads = sum(edna_reads, na.rm = TRUE), .groups = "drop") %>%
      group_by(station) %>%
      slice_max(total_reads, n = 1, with_ties = FALSE) %>%
      select(station, best_sample_type = sample_type)

    paired_pooled <- paired %>%
      left_join(best_type, by = "station") %>%
      filter(!is.na(best_sample_type), sample_type == best_sample_type) %>%
      mutate(sample_type = "Pooled (max reads)") %>%
      select(-best_sample_type) %>%
      group_by(station) %>%
      mutate(edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)) %>%
      { if (has_count)  mutate(., trawl_count_rel  = trawl_count  / sum(trawl_count,  na.rm = TRUE)) else . } %>%
      { if (has_weight) mutate(., trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) else . } %>%
      ungroup()

    cat("Pooling strategy: max_reads\n")
    cat("Stations in pooled dataset:", n_distinct(paired_pooled$station), "\n")
    best_type %>% arrange(station) %>%
      { cat("Best sample_type per station:\n"); print(as.data.frame(.)) }

  } else if (pool_strategy == "mean") {
    trawl_cols_pooled <- paired %>%
      select(station, species, any_of(c("trawl_count","trawl_weight",
                                        "trawl_count_rel","trawl_weight_rel",
                                        "taxon_group","body_type"))) %>%
      distinct()

    paired_pooled <- paired %>%
      filter(!is.na(sample_type)) %>%
      group_by(station, species) %>%
      summarise(edna_reads = mean(edna_reads, na.rm = TRUE), .groups = "drop") %>%
      left_join(trawl_cols_pooled, by = c("station","species")) %>%
      mutate(sample_type = "Pooled (mean)") %>%
      group_by(station) %>%
      mutate(edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)) %>%
      { if (has_count)  mutate(., trawl_count_rel  = trawl_count  / sum(trawl_count,  na.rm = TRUE)) else . } %>%
      { if (has_weight) mutate(., trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) else . } %>%
      ungroup()
    cat("Pooling strategy: mean\n")
    cat("Stations in pooled dataset:", n_distinct(paired_pooled$station), "\n")
  }

  # ── 3. Species overlap per station ──────────────────────────────────────────
  cat("\n### Species overlap per station\n\n")
  trawl_pa_col <- if (has_count) "trawl_count" else "trawl_weight"
  all_trawl_stations <- trawl_agg %>% distinct(station) %>% arrange(station)

  edna_summary <- paired %>%
    filter(!is.na(sample_type)) %>%
    group_by(station, sample_type) %>%
    summarise(
      n_edna = sum(edna_reads > 0),
      n_both = sum(.data[[trawl_pa_col]] > 0 & edna_reads > 0, na.rm = TRUE),
      .groups = "drop"
    )

  trawl_summary <- trawl_agg %>%
    filter(station %in% all_trawl_stations$station) %>%
    group_by(station) %>%
    summarise(n_trawl = sum(.data[[trawl_pa_col]] > 0, na.rm = TRUE), .groups = "drop")

  edna_types <- sort(unique(na.omit(paired$sample_type)))
  full_grid <- expand.grid(
    station     = all_trawl_stations$station,
    sample_type = edna_types,
    stringsAsFactors = FALSE
  )

  overlap_tbl <- full_grid %>%
    left_join(trawl_summary, by = "station") %>%
    left_join(edna_summary,  by = c("station", "sample_type")) %>%
    mutate(
      has_edna   = station %in% unique(edna_agg$station),
      n_edna_chr = if_else(!has_edna, "—", as.character(replace_na(n_edna, 0))),
      n_both_chr = if_else(!has_edna, "—", as.character(replace_na(n_both, 0))),
      pct_shared = if_else(
        !has_edna | is.na(n_both), "—",
        paste0(round(100 * n_both / pmax(n_trawl, 1), 1), "%")
      )
    ) %>%
    select(station, sample_type, n_trawl, n_edna_chr, n_both_chr, pct_shared) %>%
    arrange(station, sample_type)

  print(
    overlap_tbl %>%
      kable(
        caption   = paste0("Species detected per station [", strategy_name, "]"),
        col.names = c("Station", "eDNA sample type", "Trawl spp.", "eDNA spp.", "Shared spp.", "% Shared"),
        align     = c("l","l","c","c","c","c")
      ) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      collapse_rows(columns = 1, valign = "top")
  )

  # ── 3b. Species overlap Venn diagram (survey-wide, any eDNA sample type) ────────
  cat("\n### Species overlap Venn diagram [", strategy_name, "]\n\n")
  cat("Pooled across all stations and all eDNA sample types.",
      "A species is counted as eDNA-detected if present in ≥1 sample at ≥1 station.\n\n")

  trawl_sp_set <- trawl_agg %>%
    filter(.data[[trawl_pa_col]] > 0) %>%
    pull(species) %>% unique()

  edna_sp_set <- paired %>%
    filter(edna_reads > 0) %>%
    pull(species) %>% unique()

  trawl_only <- setdiff(trawl_sp_set, edna_sp_set)
  edna_only  <- setdiff(edna_sp_set,  trawl_sp_set)
  shared     <- intersect(trawl_sp_set, edna_sp_set)

  cat("Trawl only (", length(trawl_only), "):", paste(sort(trawl_only), collapse = ", "), "\n",
      "eDNA only  (", length(edna_only),  "):", paste(sort(edna_only),  collapse = ", "), "\n",
      "Both       (", length(shared),      "):", paste(sort(shared),      collapse = ", "), "\n\n",
      sep = "")

  p_venn <- ggvenn(
    list("Trawl" = trawl_sp_set, "Any eDNA" = edna_sp_set),
    fill_color    = c("#F08080", "#45B8AC"),
    stroke_size   = 0.5,
    set_name_size = 4,
    text_size     = 3.5
  ) +
    labs(
      title    = paste0(taxon_label, " species overlap — any eDNA sample type"),
      subtitle = paste0("[", strategy_name, "] pooled across all stations and sample types")
    ) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 13),
      plot.subtitle = element_text(hjust = 0.5, colour = "grey40")
    )
  print(p_venn)

  # ── 4a. Overall Spearman ρ ───────────────────────────────────────────────────
  cat("\n### Overall Spearman ρ [", strategy_name, "]\n\n")

  spearman_tbl <- function(df, trawl_col, metric_label) {
    df %>%
      filter(!is.na(sample_type), !is.na(.data[[trawl_col]])) %>%
      group_by(sample_type) %>%
      filter(
        sum(!is.na(.data[[trawl_col]]) & !is.na(edna_reads)) >= 3,
        var(.data[[trawl_col]], na.rm = TRUE) > 0,
        var(edna_reads,         na.rm = TRUE) > 0
      ) %>%
      summarise(
        trawl_metric = metric_label,
        n_obs   = n(),
        rho     = tryCatch(
          cor(.data[[trawl_col]], edna_reads, method = "spearman", use = "complete.obs"),
          error = function(e) NA_real_
        ),
        p_value = tryCatch(
          cor.test(.data[[trawl_col]], edna_reads, method = "spearman")$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      ) %>%
      filter(!is.na(rho))
  }

  spearman_overall <- bind_rows(
    if (has_count)  spearman_tbl(paired,        "trawl_count",  "Count"),
    if (has_weight) spearman_tbl(paired,        "trawl_weight", "Biomass"),
    if (has_count)  spearman_tbl(paired_pooled, "trawl_count",  "Count"),
    if (has_weight) spearman_tbl(paired_pooled, "trawl_weight", "Biomass")
  ) %>%
    mutate(sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ))

  print(
    spearman_overall %>%
      mutate(rho = round(rho, 3), p_value = signif(p_value, 3)) %>%
      arrange(sample_type, trawl_metric) %>%
      kable(caption = paste0("Spearman ρ [", strategy_name, "] (pooled = ", pool_strategy, ")"),
            col.names = c("Sample Type", "Trawl metric", "N obs.", "ρ", "p-value", "Sig.")) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
      collapse_rows(columns = 1, valign = "top")
  )

  # ── 4b. Station-level Spearman ρ ────────────────────────────────────────────
  spearman_station_tbl_fn <- function(df, trawl_col, metric_label) {
    df %>%
      filter(!is.na(sample_type), !is.na(.data[[trawl_col]])) %>%
      group_by(station, sample_type) %>%
      filter(
        sum(!is.na(.data[[trawl_col]]) & !is.na(edna_reads)) >= 2,
        var(.data[[trawl_col]], na.rm = TRUE) > 0,
        var(edna_reads,         na.rm = TRUE) > 0
      ) %>%
      summarise(
        trawl_metric = metric_label,
        n_spp = n(),
        rho   = tryCatch(
          cor(.data[[trawl_col]], edna_reads, method = "spearman", use = "complete.obs"),
          error = function(e) NA_real_
        ),
        .groups = "drop"
      ) %>%
      filter(!is.na(rho))
  }

  spearman_station <- bind_rows(
    if (has_count)  spearman_station_tbl_fn(paired,        "trawl_count",  "Count"),
    if (has_weight) spearman_station_tbl_fn(paired,        "trawl_weight", "Biomass"),
    if (has_count)  spearman_station_tbl_fn(paired_pooled, "trawl_count",  "Count"),
    if (has_weight) spearman_station_tbl_fn(paired_pooled, "trawl_weight", "Biomass")
  )

  if (nrow(spearman_station) > 0) {
    n_metrics  <- n_distinct(spearman_station$trawl_metric)
    all_types  <- sort(unique(spearman_station$sample_type))
    pooled_lbl <- unique(paired_pooled$sample_type)
    n_real     <- length(setdiff(all_types, pooled_lbl))
    pal_real   <- if (n_real > 0)
      RColorBrewer::brewer.pal(max(3, n_real), "Set2")[seq_len(n_real)] else character(0)
    type_colors <- setNames(c(pal_real, "#555555"),
                            c(setdiff(all_types, pooled_lbl), pooled_lbl))

    p_station <- ggplot(spearman_station,
                aes(x = station, y = rho, fill = sample_type, alpha = trawl_metric)) +
      geom_col(position = position_dodge(0.85), width = 0.8, color = "white") +
      geom_hline(yintercept = 0, linewidth = 0.4) +
      scale_fill_manual(values = type_colors) +
      scale_alpha_manual(values = c("Count" = 1, "Biomass" = 0.55), drop = FALSE) +
      labs(
        title    = paste0("Station-level Spearman ρ [", strategy_name, "]"),
        subtitle = paste0(
          if (n_metrics > 1) "Faceted by trawl metric  |  " else
            paste("Trawl metric:", unique(spearman_station$trawl_metric), " |  "),
          "Grey = pooled (", pool_strategy, ")"
        ),
        x = "Station", y = "Spearman ρ",
        fill = "Sample type", alpha = "Trawl metric"
      ) +
      theme_minimal(base_size = 13) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    if (n_metrics > 1) p_station <- p_station + facet_wrap(~trawl_metric, ncol = 1)
    print(p_station)
  }

  # ── 7. Trawl species barplots ────────────────────────────────────────────────
  cat("\n### Relative Abundance Barplots [", strategy_name, "]\n\n")

  primary_trawl_rel <- if (has_count) "trawl_count_rel" else "trawl_weight_rel"
  primary_label     <- if (has_weight) "Trawl Biomass" else "Trawl Count"
  trawl_rel_col2    <- if (has_count) "trawl_count" else "trawl_weight"

  focal_sp <- trawl_agg %>%
    filter(station %in% unique(paired$station),
           .data[[trawl_pa_col]] > 0) %>%
    pull(species) %>% unique() %>% sort()
  cat("Trawl species for barplot:", paste(focal_sp, collapse = ", "), "\n")

  # Per-station relative abundance → mean across stations.
  # Used for per-sample-type barplots (apples-to-apples with per-station eDNA rel abund).
  trawl_rel_df <- trawl_agg %>%
    filter(station %in% unique(paired$station), species %in% focal_sp) %>%
    group_by(station) %>%
    mutate(trawl_rel = .data[[trawl_rel_col2]] /
             sum(.data[[trawl_rel_col2]], na.rm = TRUE)) %>%
    ungroup() %>%
    group_by(species) %>%
    summarise(mean_trawl_rel = mean(trawl_rel, na.rm = TRUE), .groups = "drop")

  # Trawl-wide relative abundance: pool counts/weights across ALL stations first,
  # then divide once. Used for the pooled barplot so trawl values sum to 1
  # and are not inflated by species occurring at only one station.
  trawl_rel_wide_df <- trawl_agg %>%
    filter(station %in% unique(paired$station), species %in% focal_sp) %>%
    group_by(species) %>%
    summarise(total = sum(.data[[trawl_rel_col2]], na.rm = TRUE), .groups = "drop") %>%
    mutate(mean_trawl_rel = total / sum(total, na.rm = TRUE)) %>%
    select(species, mean_trawl_rel)

  edna_rel_df <- paired %>%
    filter(species %in% focal_sp, !is.na(sample_type)) %>%
    group_by(species, sample_type) %>%
    summarise(mean_edna_rel = mean(edna_rel, na.rm = TRUE), .groups = "drop")

  # Average-across-sample-types pooled eDNA:
  # For each station × species, average edna_reads across sample types, then
  # compute relative abundance within-station, then average across stations.
  edna_rel_pooled_df <- paired %>%
    filter(species %in% focal_sp, !is.na(sample_type)) %>%
    group_by(station, species) %>%
    summarise(avg_reads = mean(edna_reads, na.rm = TRUE), .groups = "drop") %>%
    group_by(station) %>%
    mutate(avg_rel = avg_reads / sum(avg_reads, na.rm = TRUE)) %>%
    ungroup() %>%
    group_by(species) %>%
    summarise(mean_edna_rel = mean(avg_rel, na.rm = TRUE), .groups = "drop")

  make_abund_barplot <- function(df_long, subtitle = NULL, not_detected_sp = character(0)) {
    method_colors <- c(
      "Trawl Biomass"    = "#F08080",
      "eDNA Read Counts" = "#45B8AC"
    )
    sp_order <- df_long %>%
      filter(method == primary_label) %>%
      arrange(desc(mean_rel_abund)) %>%
      pull(species)
    sp_order <- c(sp_order, setdiff(unique(df_long$species), sp_order))

    df_long <- df_long %>%
      mutate(species = factor(species, levels = sp_order),
             mean_rel_abund = ifelse(is.nan(mean_rel_abund), 0, mean_rel_abund))

    sp_levels <- levels(df_long$species)
    axis_labels <- if_else(
      sp_levels %in% not_detected_sp,
      paste0(sp_levels, " (x)"),
      sp_levels
    )

    not_det_df <- data.frame(
      species = factor(not_detected_sp, levels = sp_levels)
    ) %>% filter(!is.na(species))

    p <- ggplot(df_long, aes(x = species, y = mean_rel_abund,
                        fill = factor(method, levels = c(primary_label, "eDNA Read Counts")))) +
      geom_col(position = position_dodge2(padding = 0), width = 0.8) +
      scale_fill_manual(values = method_colors, name = "Method") +
      scale_y_continuous(
        labels = scales::label_number(accuracy = 0.01),
        expand = expansion(mult = c(0, 0.10))
      ) +
      scale_x_discrete(expand = expansion(add = c(0.8, 0.5)),
                       labels = axis_labels) +
      labs(
        title    = paste(taxon_label, "Species Abundance and Primer Read Count Proportions"),
        subtitle = subtitle,
        x        = "Species",
        y        = "Proportion Abundance Reads",
        caption  = if (length(not_detected_sp) > 0) "(x) = not detected by eDNA" else NULL
      ) +
      theme_bw(base_size = 12) +
      theme(
        plot.title       = element_text(face = "bold", hjust = 0.5, size = 13),
        plot.subtitle    = element_text(hjust = 0.5, colour = "grey40"),
        plot.caption     = element_text(colour = "grey50", size = 10),
        axis.text.x      = element_text(angle = 45, hjust = 1, colour = "black"),
        axis.title       = element_text(face = "bold"),
        legend.position  = "right",
        legend.title     = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        plot.margin      = margin(t = 5, r = 10, b = 5, l = 30, unit = "pt")
      )

    if (nrow(not_det_df) > 0) {
      y_max <- max(df_long$mean_rel_abund, na.rm = TRUE)
      p <- p + geom_text(
        data = not_det_df,
        aes(x = species, y = y_max * 0.04, label = "x"),
        inherit.aes = FALSE,
        colour = "grey50", size = 3.5, fontface = "bold",
        nudge_x = 0.22
      )
    }
    p
  }

  if (length(focal_sp) > 0) {
    for (stype in sort(unique(edna_rel_df$sample_type))) {
      stype_edna_rel <- edna_rel_df %>% filter(sample_type == stype)
      not_det_stype <- setdiff(focal_sp,
                               stype_edna_rel %>% filter(mean_edna_rel > 0) %>% pull(species))

      df_long <- trawl_rel_df %>%
        left_join(stype_edna_rel, by = "species") %>%
        select(species, mean_trawl_rel, mean_edna_rel) %>%
        mutate(mean_edna_rel = replace_na(mean_edna_rel, 0)) %>%
        pivot_longer(c(mean_trawl_rel, mean_edna_rel),
                     names_to = "method", values_to = "mean_rel_abund") %>%
        mutate(method = recode(method,
                               "mean_trawl_rel" = primary_label,
                               "mean_edna_rel"  = "eDNA Read Counts"))
      n_types  <- n_distinct(edna_rel_df$sample_type)
      subtitle <- paste0("[", strategy_name, "]",
                         if (n_types > 1) paste0(" | eDNA sample type: ", stype) else "")
      print(make_abund_barplot(df_long, subtitle, not_detected_sp = not_det_stype))
    }

    not_det_pooled <- setdiff(focal_sp,
                              edna_rel_pooled_df %>% filter(mean_edna_rel > 0) %>% pull(species))

    df_long_pooled <- trawl_rel_wide_df %>%
      left_join(edna_rel_pooled_df, by = "species") %>%
      mutate(mean_edna_rel = replace_na(mean_edna_rel, 0)) %>%
      pivot_longer(c(mean_trawl_rel, mean_edna_rel),
                   names_to = "method", values_to = "mean_rel_abund") %>%
      mutate(method = recode(method,
                             "mean_trawl_rel" = primary_label,
                             "mean_edna_rel"  = "eDNA Read Counts"))
    print(make_abund_barplot(
      df_long_pooled,
      subtitle        = paste0("[", strategy_name, "] Pooled eDNA & trawl-wide — average across all sample types"),
      not_detected_sp = not_det_pooled
    ))
  }

  # ── 8. Relative Read Count Export ───────────────────────────────────────────
  output_dir <- if (exists("path_analysis")) path_analysis else "."
  base_name  <- paste0("relative_reads_", tolower(gsub(" ", "_", taxon_label)),
                        "_trawl_species_", strategy_name)

  paired_stations <- unique(paired$station)

  trawl_base <- trawl_agg %>%
    filter(station %in% paired_stations) %>%
    { if (has_count)  filter(., trawl_count  > 0) else
      if (has_weight) filter(., trawl_weight > 0) else . } %>%
    group_by(station) %>%
    mutate(
      trawl_count_rel  = if (has_count)  trawl_count  / sum(trawl_count,  na.rm = TRUE) else NA_real_,
      trawl_weight_rel = if (has_weight) trawl_weight / sum(trawl_weight, na.rm = TRUE) else NA_real_
    ) %>%
    ungroup() %>%
    select(trawl_id = station, species,
           any_of(c("trawl_count", "trawl_weight",
                    "trawl_count_rel", "trawl_weight_rel")))

  trawl_species_set <- unique(trawl_base$species)

  edna_by_station <- edna_long_strat %>%
    filter(!is.na(station), !is.na(sample_type),
           station %in% paired_stations) %>%
    filter(str_to_sentence(str_squish(species)) %in%
             str_to_sentence(str_squish(trawl_species_set))) %>%
    group_by(trawl_id = station, sample_type, species) %>%
    summarise(edna_rel_read_count = mean(reads_tss, na.rm = TRUE), .groups = "drop")

  edna_wide_station <- edna_by_station %>%
    mutate(col = paste0(tolower(gsub("[ /]", "_", sample_type)), "_rel_reads")) %>%
    select(trawl_id, species, col, edna_rel_read_count) %>%
    pivot_wider(names_from = col, values_from = edna_rel_read_count, values_fill = 0)

  edna_best_station <- paired_pooled %>%
    filter(!is.na(station), !is.na(sample_type)) %>%
    filter(str_to_sentence(str_squish(species)) %in%
             str_to_sentence(str_squish(trawl_species_set))) %>%
    group_by(trawl_id = station, species) %>%
    summarise(best_rel_reads = mean(edna_rel, na.rm = TRUE), .groups = "drop")

  relreads_by_station <- trawl_base %>%
    left_join(edna_wide_station,  by = c("trawl_id", "species")) %>%
    left_join(edna_best_station,  by = c("trawl_id", "species")) %>%
    mutate(across(ends_with("_rel_reads"), ~ replace_na(.x, 0))) %>%
    arrange(trawl_id, species)

  cat("By-station export [", strategy_name, "] — rows:", nrow(relreads_by_station),
      "| stations:", n_distinct(relreads_by_station$trawl_id),
      "| species:", n_distinct(relreads_by_station$species), "\n")

  fname_station <- paste0(base_name, "_by_station.csv")
  write_csv(relreads_by_station, file.path(output_dir, fname_station))
  cat("Exported:", fname_station, "\n")

  all_trawl_st   <- unique(trawl_base$trawl_id)
  all_edna_stypes <- unique(edna_by_station$sample_type)

  trawl_wide_base <- crossing(trawl_id = all_trawl_st, species = trawl_species_set) %>%
    left_join(trawl_base, by = c("trawl_id", "species")) %>%
    mutate(across(any_of(c("trawl_count","trawl_weight","trawl_count_rel","trawl_weight_rel")),
                  ~ replace_na(.x, 0))) %>%
    group_by(species) %>%
    summarise(
      n_trawl_stations         = n(),
      n_trawl_stations_nonzero = sum(coalesce(trawl_count_rel, 0) > 0 |
                                     coalesce(trawl_weight_rel, 0) > 0),
      across(any_of(c("trawl_count","trawl_weight","trawl_count_rel","trawl_weight_rel")),
             list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE)),
             .names = "{.fn}_{.col}"),
      .groups = "drop"
    )

  edna_wide_base <- crossing(trawl_id = all_trawl_st,
                              sample_type = all_edna_stypes,
                              species = trawl_species_set) %>%
    left_join(edna_by_station, by = c("trawl_id", "sample_type", "species")) %>%
    mutate(edna_rel_read_count = replace_na(edna_rel_read_count, 0)) %>%
    group_by(sample_type, species) %>%
    summarise(
      n_edna_stations          = n(),
      n_edna_stations_detected = sum(edna_rel_read_count > 0),
      mean_edna_rel_read_count = mean(edna_rel_read_count),
      sd_edna_rel_read_count   = sd(edna_rel_read_count),
      .groups = "drop"
    )

  relreads_wide <- edna_wide_base %>%
    left_join(trawl_wide_base, by = "species") %>%
    select(sample_type, species,
           n_trawl_stations, n_trawl_stations_nonzero,
           any_of(c("mean_trawl_count", "sd_trawl_count",
                    "mean_trawl_weight", "sd_trawl_weight",
                    "mean_trawl_count_rel", "sd_trawl_count_rel",
                    "mean_trawl_weight_rel", "sd_trawl_weight_rel")),
           n_edna_stations, n_edna_stations_detected,
           mean_edna_rel_read_count, sd_edna_rel_read_count) %>%
    arrange(sample_type, desc(mean_edna_rel_read_count))

  fname_wide <- paste0(base_name, "_trawl_wide.csv")
  write_csv(relreads_wide, file.path(output_dir, fname_wide))
  cat("Exported:", fname_wide, "\n\n")

  # Preview tables
  print(
    relreads_by_station %>%
      mutate(across(where(is.numeric), ~ round(.x, 5))) %>%
      head(15) %>%
      kable(caption = paste0("By-station [", strategy_name, "] — first 15 rows")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
  )

  print(
    relreads_wide %>%
      mutate(across(where(is.numeric), ~ round(.x, 5))) %>%
      kable(caption = paste0("Trawl-wide means [", strategy_name, "]")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
  )

}  # end strategy loop
```


---

# 9. Session Info

```{r session-info}
sessionInfo()
```
