---
title: "Trawl vs. eDNA Metabarcoding: Community Survey Comparison"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

# Overview

This report harmonizes paired trawl survey and eDNA metabarcoding datasets collected at the same stations, then tests:

1. **Rank correlation** (Spearman's ρ) between trawl catch and eDNA read abundance
2. **Relative abundance correlation** across all paired station × species observations
3. **Differences among eDNA sample types** (e.g. filter type, replicate type) in their correlation with trawl data

---

# 1. Setup & Data Loading

```{r setup}
library(tidyverse)
library(data.table)
library(vegan)
library(ggplot2)
library(ggrepel)
library(ggtext)
library(ggvenn)
library(patchwork)
library(knitr)
library(kableExtra)
library(broom)

# ── USER: Update these file paths ────────────────────────────────────────────
trawl_file       <- "inputs/trawl_taxa_subset.csv"           # Long-format trawl CSV
edna_spp_files   <- c(                          # One entry per sequencing run —
  run1 = "inputs/SPECIES_counts_Trawl_MiFish12S_Run1.csv",    #   named with a short run label
  run2 = "inputs/SPECIES_counts_Trawl_MiFish12S_Run2.csv"     #   UPDATE when correct Run2 file available
)
edna_meta_file   <- "inputs/Metabarcoding_Metadata.csv"     # Sample metadata — must cover ALL runs
lookup_file      <- "inputs/species_types.csv"              # Common-name ↔ binomial lookup table

# ── USER: Update these column names to match your files ──────────────────────
# Trawl columns
trawl_station_col  <- "TowID"          # Station/site identifier
trawl_species_col  <- "Common_Name"    # Common name column in trawl
trawl_count_col    <- "Number_Caught"  # Individual counts  — set to NA if column absent
trawl_weight_col   <- "Weight_kg"      # Biomass column     — set to NA if column absent

# eDNA species table columns (same structure expected in every run file)
edna_species_col   <- "Species"        # Binomial species name column
edna_tax_col       <- "full_tax"       # Full taxonomy string column

# eDNA metadata columns
meta_sample_col    <- "ReadFile"       # Unique sample ID matching column names in eDNA tables
                                       #   NOTE: must match the sample column headers in the species table
meta_station_col   <- "TrawlID"        # Station identifier (matches trawl station IDs)
meta_type_col      <- "Sample_type"    # Column distinguishing the 3 sample types
meta_run_col       <- "RunID"          # Sequencing run identifier column in metadata
meta_primer_col    <- "Primer"         # Primer/amplicon column in metadata (set to NA if absent)

# Primer(s) to KEEP — only samples sequenced with this primer are used.
# This prevents mixing amplicons (e.g. CEPH18S + MiFish12S) when the metadata
# covers multiple primer sets. Set to NULL to keep all primers.
valid_primers <- c("MiFish12S")          # primer name exactly as it appears in the metadata
                                       # Primer column. Set to NULL to skip filter.

# Sample types to KEEP (all others — controls, mocks, NTCs — are excluded)
valid_sample_types <- c("Bottom_metaprobe", "Top_metaprobe", "Slush")

# Station ID value used for controls/blanks that should be excluded
invalid_station_values <- c("NA", "", "na")

# Lookup table columns (species_types.csv)
lookup_common_col  <- "Common_Name"    # Common name  → matches trawl species
lookup_binom_col   <- "Species"        # Binomial name → matches eDNA species
lookup_group_col   <- "Fish_Invert"    # Taxonomic group (Fish / Invertebrate)
lookup_type_col    <- "Body_Type"      # Body type (Slimy / Crunchy)

# Deduplication strategy for samples with the same station × sample_type in both runs:
# "max_reads" = keep the sample with the highest total read count (recommended)
# "run1"      = always prefer run 1
# "run2"      = always prefer run 2
dedup_strategy <- "max_reads"

# ── USER: Sample type pooling strategy for Spearman correlations ─────────────
# How to pool eDNA reads across sample types for the "Pooled" Spearman analysis:
#   "max_reads"  — per station, keep only the sample_type with the highest total
#                  read count. Preserves the strongest signal; recommended.
#   "mean"       — average edna_reads across sample types per station × species.
#                  Conservative; reduces influence of any single sample type.
pool_strategy <- "max_reads"   # ← "max_reads" or "mean"
# ─────────────────────────────────────────────────────────────────────────────

# ── USER: Taxon group filter ──────────────────────────────────────────────────
# Restrict all analyses to one taxonomic group. Options:
#   "Fish"          — keep only rows where taxon_group == "Fish"
#   "Invertebrate"  — keep only rows where taxon_group == "Invertebrate"
#   "Cephalopod"    — keep only Illex illecebrosus, Doryteuthis pealeii,
#                     and Bathypolypus arcticus (case-insensitive)
#   NULL            — no filter; use all species
taxon_filter <- "Fish"   # ← set to "Fish", "Invertebrate", "Cephalopod", or NULL

# Species included when taxon_filter == "Cephalopod"
cephalopod_species <- c("Illex illecebrosus", "Doryteuthis pealeii", "Bathypolypus arcticus")
# ─────────────────────────────────────────────────────────────────────────────
```

```{r load-data}
trawl_raw  <- read_csv(trawl_file,    show_col_types = FALSE)
meta_raw   <- read_csv(edna_meta_file, show_col_types = FALSE)
lookup_raw <- read_csv(lookup_file,    show_col_types = FALSE)

# Load each sequencing run file and tag with its run label
edna_spp_raw_list <- imap(edna_spp_files, function(path, run_label) {
  df <- read_csv(path, show_col_types = FALSE)
  df$.run <- run_label
  df
})

cat("Trawl rows:", nrow(trawl_raw), "\n")
cat("eDNA metadata rows:", nrow(meta_raw), "\n")
iwalk(edna_spp_raw_list, ~ cat("Run", .y, "—", nrow(.x), "species ×",
                                ncol(.x) - 1, "columns\n"))
cat("Lookup table rows:", nrow(lookup_raw), "\n")
```

---

# 2. Data Harmonization

## 2a. Trawl Data — Clean & Summarize

```{r trawl-clean}
trawl <- trawl_raw %>%
  rename(
    station     = all_of(trawl_station_col),
    common_name = all_of(trawl_species_col)
  ) %>%
  mutate(common_name = str_squish(common_name))

# Retain both count and biomass columns where available
has_count  <- !is.na(trawl_count_col)  && trawl_count_col  %in% names(trawl)
has_weight <- !is.na(trawl_weight_col) && trawl_weight_col %in% names(trawl)

if (!has_count && !has_weight)
  stop("Neither count nor weight column found in trawl data.")

if (has_count)  trawl <- trawl %>% rename(trawl_count  = all_of(trawl_count_col))
if (has_weight) trawl <- trawl %>% rename(trawl_weight = all_of(trawl_weight_col))

# Aggregate in case of duplicates within station × species
trawl_agg <- trawl %>%
  group_by(station, common_name) %>%
  summarise(
    across(any_of(c("trawl_count", "trawl_weight")),
           ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

cat("Trawl stations:", n_distinct(trawl_agg$station), "\n")
cat("Trawl species (common names):", n_distinct(trawl_agg$common_name), "\n")
cat("Count column present:", has_count, "| Weight column present:", has_weight, "\n")
```

## 2b. eDNA Data — Load & Combine Sequencing Runs

Each run file has the same structure (species as rows, sample IDs as columns, plus
`Species` and `full_tax` annotation columns). Runs are pivoted to long format individually
then row-bound, retaining a `seq_run` tag on every row.

```{r edna-load}
annot_cols <- c(edna_species_col, edna_tax_col)

edna_long_runs <- map(edna_spp_raw_list, function(df) {
  run_label  <- unique(df$.run)
  sample_cols <- setdiff(names(df), c(annot_cols, ".run"))

  df %>%
    filter(
      !is.na(.data[[edna_species_col]]),
      !tolower(str_squish(.data[[edna_species_col]])) %in% c("na", "")
    ) %>%
    mutate(species = str_to_sentence(str_squish(.data[[edna_species_col]]))) %>%
    select(species, full_tax = all_of(edna_tax_col), all_of(sample_cols)) %>%
    pivot_longer(
      cols      = all_of(sample_cols),
      names_to  = "sample_id",
      values_to = "reads"
    ) %>%
    filter(reads > 0) %>%
    mutate(seq_run = run_label)
})

edna_long_all <- bind_rows(edna_long_runs)

cat("Total eDNA samples across all runs:", n_distinct(edna_long_all$sample_id), "\n")
cat("eDNA species (assigned):", n_distinct(edna_long_all$species), "\n")
cat("Non-zero observations:", nrow(edna_long_all), "\n")
iwalk(edna_long_runs, ~ cat("  Run", .y, "—",
                             n_distinct(.x$sample_id), "samples,",
                             n_distinct(.x$species), "species\n"))
```

## 2c. Run Normalization & Duplicate Sample Resolution

**Why normalize?** Read depth varies between sequencing runs, so raw read counts are not
directly comparable across runs. We use Total Sum Scaling (TSS) — dividing each sample's
reads by its total reads — to convert to within-sample proportions before combining runs.
Downstream analyses use these normalized values; the heatmaps display them as proportions
rather than raw counts.

**Why deduplicate?** If the same station × sample_type combination appears in both runs,
we keep only one representative sample (using the `dedup_strategy` set above) to avoid
pseudo-replication in the correlation analyses.

```{r run-normalize-dedup}
# ── Step 1: attach metadata to get station + sample_type per sample ───────────
meta_clean_pre <- meta_raw %>%
  rename(
    sample_id   = all_of(meta_sample_col),   # ReadFile — matches species table columns
    station     = all_of(meta_station_col),   # TrawlID
    sample_type = all_of(meta_type_col)       # Sample_type
  ) %>%
  # Filter to the target primer(s) if specified
  { if (!is.null(valid_primers) && meta_primer_col %in% names(.))
      filter(., .data[[meta_primer_col]] %in% valid_primers)
    else . } %>%
  # Filter out controls, mocks, NTCs and rows with invalid station IDs
  filter(
    sample_type %in% valid_sample_types,
    !is.na(station),
    !tolower(str_squish(station)) %in% tolower(invalid_station_values)
  )

# Report primer filtering
if (!is.null(valid_primers) && meta_primer_col %in% names(meta_raw)) {
  all_primers <- sort(unique(meta_raw[[meta_primer_col]]))
  cat("Primers in metadata:", paste(all_primers, collapse = ", "), "\n")
  cat("Primers kept:", paste(valid_primers, collapse = ", "), "\n")
}
cat("Metadata rows after filtering controls + primer:", nrow(meta_clean_pre), "\n")
cat("Sample types retained:", paste(sort(unique(meta_clean_pre$sample_type)), collapse = ", "), "\n")
cat("Stations:", paste(sort(unique(meta_clean_pre$station)), collapse = ", "), "\n")

# Attach seq_run from metadata if column exists; otherwise use the run tag from the file
if (meta_run_col %in% names(meta_raw)) {
  meta_clean_pre <- meta_clean_pre %>%
    rename(seq_run_meta = all_of(meta_run_col))
} else {
  meta_clean_pre <- meta_clean_pre %>% mutate(seq_run_meta = NA_character_)
}

edna_with_meta <- edna_long_all %>%
  left_join(meta_clean_pre %>%
              select(sample_id, station, sample_type, seq_run_meta),
            by = "sample_id") %>%
  mutate(seq_run = coalesce(seq_run_meta, seq_run)) %>%
  select(-seq_run_meta) %>%
  # Drop any samples that didn't match valid metadata (controls etc.)
  filter(!is.na(station), !is.na(sample_type))

cat("\neDNA observations after joining metadata:", nrow(edna_with_meta), "\n")
cat("eDNA samples matched to valid trawl stations:",
    n_distinct(edna_with_meta$sample_id), "\n")

# ── Step 2: TSS normalization within each sample ──────────────────────────────
# IMPORTANT: compute sample_totals from edna_with_meta (after metadata join),
# NOT from edna_long_all, because seq_run has been updated via coalesce above
# and the join key must be consistent. We sum ALL species reads per sample
# (including those with 0 reads that were filtered out of edna_long_all),
# so we use the full per-sample total = sum of reads that DID make it in.
# This is a valid TSS denominator for the detected species subset.
sample_totals <- edna_with_meta %>%
  group_by(sample_id, seq_run) %>%
  summarise(total_reads = sum(reads, na.rm = TRUE), .groups = "drop")

edna_norm <- edna_with_meta %>%
  left_join(sample_totals, by = c("sample_id", "seq_run")) %>%
  mutate(reads_tss = reads / total_reads)   # proportion within sample

cat("TSS normalization check — NA total_reads:", sum(is.na(edna_norm$total_reads)), "\n")

# ── Step 3: identify duplicate station × sample_type combos across runs ──────
# A "duplicate" here means the same station × sample_type appears in BOTH runs,
# meaning multiple INDIVIDUAL samples (replicates within one run are fine and
# will be averaged later in edna_agg)
sample_index <- edna_norm %>%
  distinct(sample_id, station, sample_type, seq_run) %>%
  filter(!is.na(station), !is.na(sample_type))

# Find station × sample_type groups that span more than one sequencing run
run_per_group <- sample_index %>%
  group_by(station, sample_type) %>%
  summarise(runs_present = list(sort(unique(seq_run))),
            n_runs = n_distinct(seq_run), .groups = "drop")

dup_groups <- run_per_group %>% filter(n_runs > 1)

if (nrow(dup_groups) == 0) {
  cat("✓ No station × sample_type combinations span multiple runs.\n")
  edna_long <- edna_norm
} else {
  cat("⚠", nrow(dup_groups),
      "station × sample_type combinations have samples in >1 run:\n")
  print(dup_groups %>% select(station, sample_type, n_runs))

  # For each duplicated group, pick one run; keep all replicates within that run
  if (dedup_strategy == "max_reads") {
    # For each station × sample_type, sum reads across replicates per run, pick the run
    # with the highest total library depth
    run_depth <- sample_index %>%
      left_join(sample_totals, by = c("sample_id", "seq_run")) %>%
      group_by(station, sample_type, seq_run) %>%
      summarise(group_total = sum(total_reads, na.rm = TRUE), .groups = "drop")

    preferred_run_per_group <- run_depth %>%
      group_by(station, sample_type) %>%
      slice_max(group_total, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      select(station, sample_type, preferred_run = seq_run)

    cat("\nDeduplication strategy: keeping run with highest total reads per group.\n")
  } else {
    preferred_run_per_group <- dup_groups %>%
      mutate(preferred_run = dedup_strategy) %>%
      select(station, sample_type, preferred_run)
    cat("\nDeduplication strategy: preferring run", dedup_strategy, ".\n")
  }

  # Join preferred run back and filter — non-dup groups keep all their samples
  edna_long <- edna_norm %>%
    left_join(preferred_run_per_group, by = c("station", "sample_type")) %>%
    filter(is.na(preferred_run) | seq_run == preferred_run) %>%
    select(-preferred_run)

  cat("Samples retained after deduplication:",
      n_distinct(edna_long$sample_id), "/",
      n_distinct(edna_norm$sample_id), "\n")
}

# Summary of run composition after deduplication
run_summary <- edna_long %>%
  distinct(sample_id, station, sample_type, seq_run) %>%
  filter(!is.na(station)) %>%
  count(seq_run, name = "n_samples")
cat("\nSamples per sequencing run (post-deduplication):\n")
print(run_summary)
```

## 2d. Species Name Harmonization via Lookup Table

The trawl uses common names; the eDNA table uses binomial names. The lookup table
(`species_types.csv`) bridges them. This chunk adds the binomial key to trawl records
and flags any common names not found in the lookup.

```{r lookup-join}
# Clean lookup table
lookup <- lookup_raw %>%
  rename(
    common_name  = all_of(lookup_common_col),
    species      = all_of(lookup_binom_col),
    taxon_group  = all_of(lookup_group_col),
    body_type    = all_of(lookup_type_col)
  ) %>%
  mutate(
    common_name = str_squish(common_name),
    species     = str_to_sentence(str_squish(species))
  ) %>%
  select(common_name, species, taxon_group, body_type)

# Join binomial name onto trawl aggregated data
trawl_agg <- trawl_agg %>%
  left_join(lookup, by = "common_name")

# Report any trawl common names missing from the lookup
unmatched_trawl <- trawl_agg %>%
  filter(is.na(species)) %>%
  distinct(common_name)

if (nrow(unmatched_trawl) > 0) {
  message("⚠ These trawl common names were NOT found in the lookup table and will be excluded from correlations:")
  print(unmatched_trawl)
} else {
  cat("✓ All trawl common names matched in lookup table.\n")
}

# Report eDNA binomials not in lookup (across all runs)
all_edna_species <- map(edna_spp_raw_list, ~ {
  .x %>%
    pull(.data[[edna_species_col]]) %>%
    keep(~ !is.na(.x) && !tolower(str_squish(.x)) %in% c("na", "")) %>%
    str_to_sentence() %>% str_squish()
}) %>% unlist() %>% unique()

unmatched_edna <- setdiff(all_edna_species, lookup$species)
cat("\neDNA species not in lookup table (", length(unmatched_edna), "):",
    paste(head(unmatched_edna, 10), collapse = ", "),
    if (length(unmatched_edna) > 10) "..." else "", "\n")

# Keep only trawl records with a matched binomial
trawl_agg <- trawl_agg %>% filter(!is.na(species))

# ── Taxon group filter ────────────────────────────────────────────────────────
if (!is.null(taxon_filter)) {
  if (taxon_filter == "Cephalopod") {
    ceph_clean <- str_to_sentence(str_squish(cephalopod_species))
    trawl_agg  <- trawl_agg %>%
      filter(str_to_sentence(str_squish(species)) %in% ceph_clean)
    cat("Taxon filter: Cephalopod —", nrow(trawl_agg), "trawl records retained\n")
    cat("Species kept:", paste(sort(unique(trawl_agg$species)), collapse = ", "), "\n")
  } else {
    trawl_agg <- trawl_agg %>%
      filter(str_to_sentence(taxon_group) == str_to_sentence(taxon_filter))
    cat("Taxon filter:", taxon_filter, "—", nrow(trawl_agg), "trawl records retained\n")
    cat("Species kept:", paste(sort(unique(trawl_agg$species)), collapse = ", "), "\n")
  }
} else {
  cat("No taxon filter applied — all", nrow(trawl_agg), "trawl records retained\n")
}

# Derive the group label used in plot titles
taxon_label <- if (is.null(taxon_filter)) "All species" else taxon_filter
```

## 2e. eDNA — Attach Metadata & Summarize by Station × Type

```{r edna-meta}
# Station and sample_type are already attached from the normalization step.
# Re-use meta_clean_pre for any remaining metadata columns needed.
meta_clean <- meta_clean_pre   # already renamed in 2c

# Warn about samples in eDNA tables with no metadata match
missing_meta <- setdiff(unique(edna_long_all$sample_id), meta_clean$sample_id)
if (length(missing_meta) > 0)
  warning("These sample IDs are in the eDNA table but not in metadata: ",
          paste(missing_meta, collapse = ", "))

# Summarize TSS-normalized reads per station × sample_type × species
# (using reads_tss so cross-run values are on the same 0–1 proportion scale)
edna_agg <- edna_long %>%
  filter(!is.na(station), !is.na(sample_type)) %>%
  group_by(station, sample_type, species) %>%
  summarise(edna_reads = sum(reads_tss, na.rm = TRUE), .groups = "drop")

cat("eDNA sample types:", paste(sort(unique(edna_agg$sample_type)), collapse = ", "), "\n")
cat("eDNA stations:", n_distinct(edna_agg$station), "\n")
```

## 2f. Join Trawl & eDNA

```{r join}
shared_stations <- intersect(unique(trawl_agg$station), unique(edna_agg$station))
cat("Paired stations (trawl ∩ eDNA):", length(shared_stations), "\n")

# Full join on binomial species name within paired stations
paired <- trawl_agg %>%
  filter(station %in% shared_stations) %>%
  full_join(
    edna_agg %>% filter(station %in% shared_stations),
    by = c("station", "species")
  ) %>%
  # Fill zeros for species absent from one method at a given station
  mutate(
    across(any_of(c("trawl_count", "trawl_weight")), ~ replace_na(.x, 0)),
    edna_reads = replace_na(edna_reads, 0)
  ) %>%
  # Carry lookup metadata forward (taxon_group, body_type)
  left_join(lookup %>% select(species, taxon_group, body_type), by = "species")

# Relative abundances within each station × sample_type (computed per metric)
paired <- paired %>%
  group_by(station, sample_type) %>%
  mutate(
    edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)
  ) %>%
  ungroup()

if (has_count) {
  paired <- paired %>%
    group_by(station, sample_type) %>%
    mutate(trawl_count_rel = trawl_count / sum(trawl_count, na.rm = TRUE)) %>%
    ungroup()
}
if (has_weight) {
  paired <- paired %>%
    group_by(station, sample_type) %>%
    mutate(trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) %>%
    ungroup()
}

both_detected <- paired %>% filter(edna_reads > 0)
if (has_count)  both_detected <- both_detected %>% filter(trawl_count  > 0)
if (has_weight) both_detected <- both_detected %>% filter(trawl_weight > 0)

cat("Species × station pairs detected by both methods:", nrow(both_detected), "\n")

# ── Pooled eDNA: collapse sample_types within each station ───────────────────
# Used for the pooled Spearman analyses in Section 4.
if (pool_strategy == "max_reads") {
  # Per station, find the sample_type with the highest total read count.
  # Use edna_agg (pre-join) as the source — paired has NAs for stations
  # where species were only in trawl, which inflates or drops totals.
  best_type <- edna_agg %>%
    group_by(station, sample_type) %>%
    summarise(total_reads = sum(edna_reads, na.rm = TRUE), .groups = "drop") %>%
    group_by(station) %>%
    slice_max(total_reads, n = 1, with_ties = FALSE) %>%
    select(station, best_sample_type = sample_type)

  paired_pooled <- paired %>%
    left_join(best_type, by = "station") %>%
    filter(!is.na(best_sample_type), sample_type == best_sample_type) %>%
    mutate(sample_type = "Pooled (max reads)") %>%
    select(-best_sample_type) %>%
    # Recompute relative abundances within each station using only the
    # selected sample_type's rows (inherited edna_rel was computed across
    # all types before pooling and may be stale / NaN)
    group_by(station) %>%
    mutate(
      edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)
    ) %>%
    { if (has_count)  mutate(., trawl_count_rel  = trawl_count  / sum(trawl_count,  na.rm = TRUE)) else . } %>%
    { if (has_weight) mutate(., trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) else . } %>%
    ungroup()

  cat("Pooling strategy: max_reads\n")
  cat("Stations in pooled dataset:", n_distinct(paired_pooled$station), "\n")
  best_type %>% arrange(station) %>%
    { cat("Best sample_type per station:\n"); print(as.data.frame(.)) }

} else if (pool_strategy == "mean") {
  # Average edna_reads across sample types per station × species
  trawl_cols_pooled <- paired %>%
    select(station, species, any_of(c("trawl_count","trawl_weight",
                                      "trawl_count_rel","trawl_weight_rel",
                                      "taxon_group","body_type"))) %>%
    distinct()

  paired_pooled <- paired %>%
    filter(!is.na(sample_type)) %>%
    group_by(station, species) %>%
    summarise(edna_reads = mean(edna_reads, na.rm = TRUE), .groups = "drop") %>%
    left_join(trawl_cols_pooled, by = c("station","species")) %>%
    mutate(sample_type = "Pooled (mean)") %>%
    group_by(station) %>%
    mutate(
      edna_rel = edna_reads / sum(edna_reads, na.rm = TRUE)
    ) %>%
    { if (has_count)  mutate(., trawl_count_rel  = trawl_count  / sum(trawl_count,  na.rm = TRUE)) else . } %>%
    { if (has_weight) mutate(., trawl_weight_rel = trawl_weight / sum(trawl_weight, na.rm = TRUE)) else . } %>%
    ungroup()
  cat("Pooling strategy: mean\n")
  cat("Stations in pooled dataset:", n_distinct(paired_pooled$station), "\n")
}
# ─────────────────────────────────────────────────────────────────────────────
```

---

# 3. Exploratory Summaries

## Species overlap per station

All trawl stations are shown regardless of whether a paired eDNA sample exists. Stations
with no eDNA data show `—` in the eDNA columns.

```{r species-overlap}
trawl_pa_col <- if (has_count) "trawl_count" else "trawl_weight"

# All trawl stations (the authoritative list)
all_trawl_stations <- trawl_agg %>% distinct(station) %>% arrange(station)

# eDNA summary per station × sample_type
edna_summary <- paired %>%
  filter(!is.na(sample_type)) %>%
  group_by(station, sample_type) %>%
  summarise(
    n_edna = sum(edna_reads > 0),
    n_both = sum(.data[[trawl_pa_col]] > 0 & edna_reads > 0, na.rm = TRUE),
    .groups = "drop"
  )

# Trawl summary per station (independent of eDNA sample_type)
trawl_summary <- trawl_agg %>%
  filter(station %in% all_trawl_stations$station) %>%
  group_by(station) %>%
  summarise(n_trawl = sum(.data[[trawl_pa_col]] > 0, na.rm = TRUE), .groups = "drop")

# Full cross of all trawl stations × eDNA sample types that exist in the data
edna_types <- sort(unique(na.omit(paired$sample_type)))
full_grid <- expand.grid(
  station     = all_trawl_stations$station,
  sample_type = edna_types,
  stringsAsFactors = FALSE
)

overlap_tbl <- full_grid %>%
  left_join(trawl_summary, by = "station") %>%
  left_join(edna_summary,  by = c("station", "sample_type")) %>%
  mutate(
    has_edna   = station %in% unique(edna_agg$station),
    n_edna_chr = if_else(!has_edna, "—", as.character(replace_na(n_edna, 0))),
    n_both_chr = if_else(!has_edna, "—", as.character(replace_na(n_both, 0))),
    pct_shared = if_else(
      !has_edna | is.na(n_both), "—",
      paste0(round(100 * n_both / pmax(n_trawl, 1), 1), "%")
    )
  ) %>%
  select(station, sample_type, n_trawl, n_edna_chr, n_both_chr, pct_shared) %>%
  arrange(station, sample_type)

overlap_tbl %>%
  kable(
    caption   = "Species detected per station by each method (all trawl stations shown)",
    col.names = c("Station", "eDNA sample type", "Trawl spp.", "eDNA spp.", "Shared spp.", "% Shared"),
    align     = c("l","l","c","c","c","c")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

## Species overlap Venn diagrams

Four pairwise comparisons between trawl and each eDNA sample type (plus all eDNA types
pooled). Species sets are based on detections across **all paired stations**. Shared
species are listed below each diagram.

```{r venn-diagrams, fig.height=4.5, fig.width=5.5}
# Species sets — trawl and each eDNA type, restricted to shared stations
trawl_spp_set <- trawl_agg %>%
  filter(station %in% shared_stations, .data[[trawl_pa_col]] > 0) %>%
  pull(species) %>% unique() %>% sort()

edna_spp_by_type <- edna_agg %>%
  filter(station %in% shared_stations, edna_reads > 0) %>%
  group_by(sample_type) %>%
  summarise(species = list(sort(unique(species))), .groups = "drop") %>%
  deframe()   # named list: sample_type -> character vector of species

# Add pooled "All eDNA" set
edna_spp_by_type[["All eDNA"]] <- edna_agg %>%
  filter(station %in% shared_stations, edna_reads > 0) %>%
  pull(species) %>% unique() %>% sort()

# Helper: draw one Venn + print shared species beneath
venn_with_list <- function(edna_type, edna_spp) {
  shared <- sort(intersect(trawl_spp_set, edna_spp))
  only_trawl <- sort(setdiff(trawl_spp_set, edna_spp))
  only_edna  <- sort(setdiff(edna_spp, trawl_spp_set))

  venn_sets <- list(Trawl = trawl_spp_set, eDNA = edna_spp)
  names(venn_sets)[2] <- edna_type

  p <- ggvenn(
    venn_sets,
    fill_color   = c("#4e79a7", "#f28e2b"),
    fill_alpha   = 0.45,
    stroke_size  = 0.6,
    text_size    = 4,
    set_name_size = 4.5
  ) +
    labs(title = paste("Trawl vs.", edna_type)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13))

  print(p)

  # Species lists beneath the diagram
  cat("\n**Shared species (n =", length(shared), "):**\n")
  if (length(shared) > 0) {
    cat(paste0("- *", shared, "*", collapse = "\n"), "\n")
  } else {
    cat("- None\n")
  }

  cat("\n**Trawl only (n =", length(only_trawl), "):**",
      if (length(only_trawl) > 0) paste0(" ", paste(only_trawl, collapse = ", ")) else " None",
      "\n")
  cat("\n**", edna_type, "only (n =", length(only_edna), "):**",
      if (length(only_edna) > 0) paste0(" ", paste(only_edna, collapse = ", ")) else " None",
      "\n\n---\n\n")
}

# Draw each comparison
for (etype in c(sort(names(edna_spp_by_type)[names(edna_spp_by_type) != "All eDNA"]),
                "All eDNA")) {
  venn_with_list(etype, edna_spp_by_type[[etype]])
}
```

## Per-species heatmaps: relative abundance × station × method

Each panel shows one species detected by **both** trawl and eDNA across paired stations.
Rows = all trawl stations; columns = Trawl + eDNA sample types. Colour encodes
**trawl relative abundance** (Trawl column) or **raw eDNA read counts** (eDNA columns),
scaled 0–1 within each column. Grey tiles = stations with no eDNA sample.

```{r species-heatmap, fig.height=5}
# ── Build the heatmap data ────────────────────────────────────────────────────
trawl_heat <- trawl_agg %>%
  filter(station %in% all_trawl_stations$station) %>%
  group_by(station) %>%
  mutate(trawl_rel_abund = .data[[trawl_pa_col]] / sum(.data[[trawl_pa_col]], na.rm = TRUE)) %>%
  ungroup() %>%
  select(station, species, value = trawl_rel_abund) %>%
  mutate(col_group = "Trawl")

edna_heat <- edna_agg %>%
  filter(station %in% all_trawl_stations$station) %>%
  rename(value = edna_reads) %>%
  mutate(col_group = sample_type) %>%
  select(station, species, value, col_group)

col_levels <- c("Trawl", sort(unique(edna_heat$col_group)))

# ── Restrict to species detected by BOTH trawl and eDNA ──────────────────────
trawl_detected <- trawl_agg %>%
  filter(station %in% shared_stations, .data[[trawl_pa_col]] > 0) %>%
  pull(species) %>% unique()

edna_detected <- edna_agg %>%
  filter(station %in% shared_stations, edna_reads > 0) %>%
  pull(species) %>% unique()

both_species <- sort(intersect(trawl_detected, edna_detected))

if (length(both_species) == 0) {
  cat("No species detected by both trawl and eDNA in paired stations.\n")
} else {
  cat("Species detected by both methods:", paste(both_species, collapse = ", "), "\n")

  heat_grid <- expand.grid(
    station   = all_trawl_stations$station,
    species   = both_species,
    col_group = col_levels,
    stringsAsFactors = FALSE
  )

  heat_data <- heat_grid %>%
    left_join(bind_rows(trawl_heat, edna_heat),
              by = c("station", "species", "col_group")) %>%
    mutate(
      has_edna  = station %in% unique(edna_agg$station),
      is_trawl  = col_group == "Trawl",
      missing   = !is_trawl & !has_edna,
      value     = if_else(missing, NA_real_, replace_na(value, 0)),
      col_group = factor(col_group, levels = col_levels),
      station   = factor(station, levels = rev(all_trawl_stations$station))
    )

  for (sp in both_species) {
    sp_data <- heat_data %>% filter(species == sp)

    sp_data <- sp_data %>%
      group_by(col_group) %>%
      mutate(value_scaled = case_when(
        is.na(value)                  ~ NA_real_,
        max(value, na.rm = TRUE) == 0 ~ 0,
        TRUE ~ value / max(value, na.rm = TRUE)
      )) %>%
      ungroup()

    p <- ggplot(sp_data, aes(x = col_group, y = station, fill = value_scaled)) +
      geom_tile(color = "white", linewidth = 0.5) +
      geom_text(
        aes(label = case_when(
          missing      ~ "",
          is.na(value) ~ "0",
          is_trawl     ~ scales::percent(value, accuracy = 0.1),
          TRUE         ~ scales::percent(value, accuracy = 0.01)
        )),
        size = 2.8, color = "grey20"
      ) +
      scale_fill_gradientn(
        colours  = c("#f0f7ff", "#6baed6", "#08519c"),
        na.value = "#e8e8e8",
        limits   = c(0, 1),
        labels   = scales::percent,
        name     = "Scaled value\n(within column)"
      ) +
      scale_x_discrete(position = "top") +
      labs(
        title    = paste0("*", sp, "*"),
        subtitle = "Trawl = relative abundance  |  eDNA = TSS-normalized proportion  |  Grey = no eDNA sample",
        x = NULL, y = "Station"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        plot.title       = ggtext::element_markdown(face = "italic", size = 12),
        axis.text.x      = element_text(angle = 30, hjust = 0, face = "bold"),
        panel.grid       = element_blank(),
        legend.key.width = unit(1.2, "cm")
      )

    print(p)
  }
}
```

---

# 4. Rank Correlation (Spearman's ρ)

## 4a. Overall rank correlation per sample type

Spearman's ρ is calculated across all station × species pairs within each eDNA sample type,
separately for **count** and **biomass** where both are available. This lets you directly
compare which trawl metric correlates more strongly with eDNA read abundance.

```{r spearman-overall}
# Helper: compute Spearman ρ + p for one trawl metric
spearman_tbl <- function(df, trawl_col, metric_label) {
  df %>%
    filter(!is.na(sample_type), !is.na(.data[[trawl_col]])) %>%
    group_by(sample_type) %>%
    filter(
      sum(!is.na(.data[[trawl_col]]) & !is.na(edna_reads)) >= 3,
      var(.data[[trawl_col]], na.rm = TRUE) > 0,
      var(edna_reads,         na.rm = TRUE) > 0
    ) %>%
    summarise(
      trawl_metric = metric_label,
      n_obs   = n(),
      rho     = tryCatch(
        cor(.data[[trawl_col]], edna_reads, method = "spearman", use = "complete.obs"),
        error = function(e) NA_real_
      ),
      p_value = tryCatch(
        cor.test(.data[[trawl_col]], edna_reads, method = "spearman")$p.value,
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) %>%
    filter(!is.na(rho))
}

spearman_overall <- bind_rows(
  if (has_count)  spearman_tbl(paired,        "trawl_count",  "Count"),
  if (has_weight) spearman_tbl(paired,        "trawl_weight", "Biomass"),
  if (has_count)  spearman_tbl(paired_pooled, "trawl_count",  "Count"),
  if (has_weight) spearman_tbl(paired_pooled, "trawl_weight", "Biomass")
) %>%
  mutate(sig = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    TRUE            ~ "ns"
  ))

spearman_overall %>%
  mutate(rho = round(rho, 3), p_value = signif(p_value, 3)) %>%
  arrange(sample_type, trawl_metric) %>%
  kable(caption = paste("Spearman ρ: trawl vs eDNA reads, by sample type (pooled =",
                        pool_strategy, ")"),
        col.names = c("Sample Type", "Trawl metric", "N obs.", "ρ", "p-value", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

## 4b. Station-level Spearman ρ — Count vs Biomass

```{r spearman-station}
# Helper: station-level ρ for one trawl metric
# Guards against groups with too few non-zero observations or zero variance
spearman_station_tbl <- function(df, trawl_col, metric_label) {
  df %>%
    filter(!is.na(sample_type), !is.na(.data[[trawl_col]])) %>%
    group_by(station, sample_type) %>%
    filter(
      sum(!is.na(.data[[trawl_col]]) & !is.na(edna_reads)) >= 2,
      var(.data[[trawl_col]], na.rm = TRUE) > 0,
      var(edna_reads,         na.rm = TRUE) > 0
    ) %>%
    summarise(
      trawl_metric = metric_label,
      n_spp = n(),
      rho   = tryCatch(
        cor(.data[[trawl_col]], edna_reads, method = "spearman", use = "complete.obs"),
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) %>%
    filter(!is.na(rho))
}

spearman_station <- bind_rows(
  if (has_count)  spearman_station_tbl(paired,        "trawl_count",  "Count"),
  if (has_weight) spearman_station_tbl(paired,        "trawl_weight", "Biomass"),
  if (has_count)  spearman_station_tbl(paired_pooled, "trawl_count",  "Count"),
  if (has_weight) spearman_station_tbl(paired_pooled, "trawl_weight", "Biomass")
)

if (nrow(spearman_station) == 0) {
  cat("No station × sample_type groups had sufficient variance for station-level ρ.\n")
} else {
  n_metrics  <- n_distinct(spearman_station$trawl_metric)
  all_types  <- sort(unique(spearman_station$sample_type))
  pooled_lbl <- unique(paired_pooled$sample_type)

  # Give the pooled level a distinct colour (dark grey) so it stands out
  n_real   <- length(setdiff(all_types, pooled_lbl))
  pal_real <- if (n_real > 0)
    RColorBrewer::brewer.pal(max(3, n_real), "Set2")[seq_len(n_real)] else character(0)
  type_colors <- setNames(c(pal_real, "#555555"),
                          c(setdiff(all_types, pooled_lbl), pooled_lbl))

  p <- ggplot(spearman_station,
              aes(x = station, y = rho, fill = sample_type, alpha = trawl_metric)) +
    geom_col(position = position_dodge(0.85), width = 0.8, color = "white") +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    scale_fill_manual(values = type_colors) +
    scale_alpha_manual(values = c("Count" = 1, "Biomass" = 0.55), drop = FALSE) +
    labs(
      title    = "Station-level Spearman ρ: Trawl vs eDNA",
      subtitle = paste0(
        if (n_metrics > 1) "Faceted by trawl metric  |  " else
          paste("Trawl metric:", unique(spearman_station$trawl_metric), " |  "),
        "Grey = pooled (", pool_strategy, ")"
      ),
      x = "Station", y = "Spearman ρ",
      fill = "Sample type", alpha = "Trawl metric"
    ) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  if (n_metrics > 1) p <- p + facet_wrap(~trawl_metric, ncol = 1)

  print(p)
}
```

---

# 5. Relative Abundance Correlation

## 5a. Scatter plots per sample type — Count vs Biomass

```{r relabund-scatter, fig.height=7}
rel_long <- paired %>%
  filter(!is.na(sample_type)) %>%
  select(station, species, sample_type, edna_rel,
         any_of(c("trawl_count_rel", "trawl_weight_rel"))) %>%
  pivot_longer(
    cols      = any_of(c("trawl_count_rel", "trawl_weight_rel")),
    names_to  = "trawl_metric",
    values_to = "trawl_rel"
  ) %>%
  filter(!is.na(trawl_rel), !is.na(edna_rel),
         (trawl_rel > 0 | edna_rel > 0)) %>%
  mutate(trawl_metric = recode(trawl_metric,
                               "trawl_count_rel"  = "Count",
                               "trawl_weight_rel" = "Biomass"))

if (nrow(rel_long) == 0) {
  cat("No paired relative abundance data to plot.\n")
} else {
  metrics <- unique(rel_long$trawl_metric)
  types   <- sort(unique(rel_long$sample_type))
  colors  <- setNames(RColorBrewer::brewer.pal(max(3, length(types)), "Set2")[seq_along(types)],
                      types)

  # Build one plot per metric × type combination, only where data exist
  plot_list <- list()
  for (m in metrics) {
    for (t in types) {
      sub <- rel_long %>% filter(trawl_metric == m, sample_type == t)
      if (nrow(sub) == 0) next

      p_sub <- ggplot(sub, aes(x = trawl_rel, y = edna_rel)) +
        geom_point(color = colors[t], alpha = 0.6, size = 1.8)

      if (nrow(sub) >= 3) {
        p_sub <- p_sub +
          geom_smooth(method = "lm", se = FALSE, color = colors[t], linewidth = 0.8)
      }

      p_sub <- p_sub +
        scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
        scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
        labs(title = paste0(t, "\n(", m, ")"),
             x = "Trawl rel. abund.", y = "eDNA rel. abund.") +
        theme_bw(base_size = 10) +
        theme(plot.title = element_text(size = 9))

      plot_list[[paste(m, t, sep = "_")]] <- p_sub
    }
  }

  if (length(plot_list) > 0) {
    wrap_plots(plot_list, ncol = length(types)) +
      plot_annotation(
        title    = "Relative Abundance: Trawl vs eDNA",
        subtitle = "Smooth line shown only where ≥3 paired observations exist"
      )
  }
}
```

## 5b. Log-transformed Pearson r — Count vs Biomass

```{r relabund-log}
log_cor <- bind_rows(
  if (has_count) {
    paired %>%
      filter(!is.na(sample_type)) %>%
      group_by(sample_type) %>%
      filter(
        sum(!is.na(trawl_count) & !is.na(edna_reads)) >= 3,
        var(log1p(trawl_count), na.rm = TRUE) > 0,
        var(log1p(edna_reads),  na.rm = TRUE) > 0
      ) %>%
      summarise(
        trawl_metric = "Count",
        pearson_r = tryCatch(
          cor(log1p(trawl_count), log1p(edna_reads), use = "complete.obs"),
          error = function(e) NA_real_
        ),
        p_value = tryCatch(
          cor.test(log1p(trawl_count), log1p(edna_reads))$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      ) %>% filter(!is.na(pearson_r))
  },
  if (has_weight) {
    paired %>%
      filter(!is.na(sample_type)) %>%
      group_by(sample_type) %>%
      filter(
        sum(!is.na(trawl_weight) & !is.na(edna_reads)) >= 3,
        var(log1p(trawl_weight), na.rm = TRUE) > 0,
        var(log1p(edna_reads),   na.rm = TRUE) > 0
      ) %>%
      summarise(
        trawl_metric = "Biomass",
        pearson_r = tryCatch(
          cor(log1p(trawl_weight), log1p(edna_reads), use = "complete.obs"),
          error = function(e) NA_real_
        ),
        p_value = tryCatch(
          cor.test(log1p(trawl_weight), log1p(edna_reads))$p.value,
          error = function(e) NA_real_
        ),
        .groups = "drop"
      ) %>% filter(!is.na(pearson_r))
  }
) %>%
  mutate(sig = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    TRUE ~ "ns"
  ))

log_cor %>%
  mutate(pearson_r = round(pearson_r, 3), p_value = signif(p_value, 3)) %>%
  arrange(sample_type, trawl_metric) %>%
  kable(caption = "Pearson r on log1p(abundance): Count vs Biomass, by sample type",
        col.names = c("Sample Type", "Trawl metric", "Pearson r", "p-value", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

---

# 6. Differences Among eDNA Sample Types

## 6a. Kruskal-Wallis + Dunn post-hoc test on ρ values

```{r kruskal-type}
n_types_kw <- n_distinct(spearman_station$sample_type)

if (nrow(spearman_station) == 0 || n_types_kw < 2) {
  cat("Kruskal-Wallis test requires ≥2 sample types with station-level ρ values.\n",
      "Only", n_types_kw, "type(s) available after variance filtering — skipping.\n")
} else {
  # Test whether station-level ρ differs by sample type
  kw_result <- kruskal.test(rho ~ sample_type, data = spearman_station)
  kw_tidy   <- tidy(kw_result)

  cat("Kruskal-Wallis test: H =", round(kw_tidy$statistic, 3),
      ", df =", kw_tidy$parameter,
      ", p =", signif(kw_tidy$p.value, 3), "\n")

  # Pairwise Wilcoxon (Dunn-like) if K-W is significant
  if (!is.na(kw_tidy$p.value) && kw_tidy$p.value < 0.05 && n_types_kw >= 2) {
    pairwise_result <- pairwise.wilcox.test(
      spearman_station$rho,
      spearman_station$sample_type,
      p.adjust.method = "BH"
    )
    cat("\nPairwise Wilcoxon (BH-adjusted p-values):\n")
    print(pairwise_result$p.value)
  } else {
    cat("No significant difference among sample types, skipping post-hoc.\n")
  }
}
```

## 6b. Boxplot of station-level ρ by sample type

```{r rho-boxplot, fig.height=5}
if (nrow(spearman_station) == 0) {
  cat("No station-level ρ values available to plot.\n")
} else {
  n_metrics <- n_distinct(spearman_station$trawl_metric)

  p <- ggplot(spearman_station, aes(x = sample_type, y = rho, fill = sample_type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.5) +
    geom_jitter(width = 0.15, size = 2.2, alpha = 0.8, color = "grey30") +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(
      title    = "Station-level Spearman ρ by eDNA Sample Type",
      subtitle = if (n_metrics > 1) "Faceted by trawl metric" else
                   paste("Trawl metric:", unique(spearman_station$trawl_metric)),
      x = "Sample type", y = "Spearman ρ (trawl vs eDNA)"
    ) +
    theme_bw(base_size = 13)

  if (n_metrics > 1) p <- p + facet_wrap(~trawl_metric, ncol = 2)

  print(p)
}
```

## 6c. NMDS ordination — do sample types cluster separately?

```{r nmds, fig.height=6}
# Build community matrix: station × species, summed reads, per sample type
nmds_list <- edna_agg %>%
  filter(station %in% shared_stations) %>%
  group_by(station, sample_type) %>%
  pivot_wider(names_from = species, values_from = edna_reads, values_fill = 0) %>%
  ungroup()

nmds_meta <- nmds_list %>% select(station, sample_type)
nmds_mat  <- nmds_list %>% select(-station, -sample_type) %>% as.matrix()

# Remove all-zero rows/cols
keep_rows <- rowSums(nmds_mat) > 0
keep_cols <- colSums(nmds_mat) > 0
nmds_mat  <- nmds_mat[keep_rows, keep_cols]
nmds_meta <- nmds_meta[keep_rows, ]

if (nrow(nmds_mat) >= 3) {
  set.seed(42)
  ord <- metaMDS(nmds_mat, distance = "bray", k = 2, trymax = 100, trace = FALSE)

  scores_df <- as.data.frame(scores(ord, "sites")) %>%
    bind_cols(nmds_meta)

  ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = sample_type, shape = sample_type)) +
    geom_point(size = 3.5, alpha = 0.85) +
    geom_text_repel(aes(label = station), size = 3, max.overlaps = 10) +
    stat_ellipse(aes(group = sample_type), type = "t", level = 0.75, linewidth = 0.6) +
    scale_color_brewer(palette = "Set2") +
    labs(
      title = "NMDS Ordination of eDNA Communities",
      subtitle = paste0("Bray-Curtis dissimilarity | stress = ", round(ord$stress, 3)),
      color = "Sample type", shape = "Sample type"
    ) +
    theme_bw(base_size = 13)
} else {
  cat("Insufficient observations for NMDS.\n")
}
```

## 6d. PERMANOVA — do sample types differ in community composition?

```{r permanova}
if (nrow(nmds_mat) >= 3) {
  perm_result <- adonis2(
    nmds_mat ~ sample_type + station,
    data    = nmds_meta,
    method  = "bray",
    permutations = 999
  )
  print(perm_result)
} else {
  cat("Insufficient observations for PERMANOVA.\n")
}
```

---

# 7. Trawl Species: Relative Abundance Comparison

Mean relative abundance (trawl) and mean relative eDNA read count side by side for all
species **detected by trawl** across paired stations. Species detected only by eDNA are
excluded so the x-axis reflects the trawl community composition.

```{r top-species, fig.height=7}
# Choose the primary trawl relative abundance metric
primary_trawl_rel <- if (has_count) "trawl_count_rel" else "trawl_weight_rel"
primary_label     <- if (has_count) "Trawl biomass"   else "Trawl biomass"
trawl_pa_col      <- if (has_count) "trawl_count"     else "trawl_weight"

# Species detected by trawl in paired stations (eDNA-only excluded)
focal_sp <- trawl_agg %>%
  filter(station %in% unique(paired$station),
         .data[[trawl_pa_col]] > 0) %>%
  pull(species) %>% unique() %>% sort()
cat("Trawl species for barplot:", paste(focal_sp, collapse = ", "), "\n")

# ── Trawl relative abundances (from trawl_agg directly, not paired) ──────────
# Use trawl_agg so we don't inherit NaN from the full-join's zero-fill logic.
trawl_rel_col <- if (has_count) "trawl_count" else "trawl_weight"
trawl_rel_df <- trawl_agg %>%
  filter(station %in% unique(paired$station), species %in% focal_sp) %>%
  group_by(station) %>%
  mutate(trawl_rel = .data[[trawl_rel_col]] /
           sum(.data[[trawl_rel_col]], na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(species) %>%
  summarise(mean_trawl_rel = mean(trawl_rel, na.rm = TRUE), .groups = "drop")

# ── eDNA relative abundances per sample_type (from paired) ───────────────────
edna_rel_df <- paired %>%
  filter(species %in% focal_sp, !is.na(sample_type)) %>%
  group_by(species, sample_type) %>%
  summarise(mean_edna_rel = mean(edna_rel, na.rm = TRUE), .groups = "drop")

# ── eDNA relative abundances for pooled ──────────────────────────────────────
edna_rel_pooled_df <- paired_pooled %>%
  filter(species %in% focal_sp) %>%
  group_by(species) %>%
  summarise(mean_edna_rel = mean(edna_rel, na.rm = TRUE), .groups = "drop")

# ── Shared plot helper ────────────────────────────────────────────────────────
make_abund_barplot <- function(df_long, subtitle = NULL) {
  method_colors <- c(
    "Trawl biomass"    = "#F08080",
    "eDNA Read Counts" = "#45B8AC"
  )
  sp_order <- df_long %>%
    filter(method == primary_label) %>%
    arrange(desc(mean_rel_abund)) %>%
    pull(species)
  sp_order <- c(sp_order, setdiff(unique(df_long$species), sp_order))

  df_long <- df_long %>%
    mutate(species = factor(species, levels = sp_order),
           mean_rel_abund = ifelse(is.nan(mean_rel_abund), 0, mean_rel_abund))

  ggplot(df_long, aes(x = species, y = mean_rel_abund, fill = method)) +
    geom_col(position = position_dodge(0.75), width = 0.65) +
    scale_fill_manual(values = method_colors, name = "Method") +
    scale_y_continuous(
      labels = scales::label_number(accuracy = 0.01),
      expand = expansion(mult = c(0, 0.08))
    ) +
    labs(
      title    = paste(taxon_label, "Species Abundance and Primer Read Count Proportions"),
      subtitle = subtitle,
      x        = "Species",
      y        = "Proportion Abundance Reads"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", hjust = 0.5, size = 13),
      plot.subtitle    = element_text(hjust = 0.5, colour = "grey40"),
      axis.text.x      = element_text(angle = 45, hjust = 1, colour = "purple"),
      axis.title       = element_text(face = "bold"),
      legend.position  = "right",
      legend.title     = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}
# ─────────────────────────────────────────────────────────────────────────────

if (length(focal_sp) == 0) {
  cat("No trawl species found in paired stations.\n")
} else {
  # One chart per sample_type
  for (stype in sort(unique(edna_rel_df$sample_type))) {
    df_long <- trawl_rel_df %>%
      left_join(edna_rel_df %>% filter(sample_type == stype),
                by = "species") %>%
      select(species, mean_trawl_rel, mean_edna_rel) %>%
      mutate(mean_edna_rel = replace_na(mean_edna_rel, 0)) %>%
      pivot_longer(c(mean_trawl_rel, mean_edna_rel),
                   names_to = "method", values_to = "mean_rel_abund") %>%
      mutate(method = recode(method,
                             "mean_trawl_rel" = primary_label,
                             "mean_edna_rel"  = "eDNA Read Counts"))

    n_types  <- n_distinct(edna_rel_df$sample_type)
    subtitle <- if (n_types > 1) paste("eDNA sample type:", stype) else NULL
    print(make_abund_barplot(df_long, subtitle))
  }

  # Pooled chart
  df_long_pooled <- trawl_rel_df %>%
    left_join(edna_rel_pooled_df, by = "species") %>%
    mutate(mean_edna_rel = replace_na(mean_edna_rel, 0)) %>%
    pivot_longer(c(mean_trawl_rel, mean_edna_rel),
                 names_to = "method", values_to = "mean_rel_abund") %>%
    mutate(method = recode(method,
                           "mean_trawl_rel" = primary_label,
                           "mean_edna_rel"  = "eDNA Read Counts"))

  print(make_abund_barplot(
    df_long_pooled,
    subtitle = paste0("Pooled eDNA — best sample type per station (", pool_strategy, ")")
  ))
}
```

---

# 8. Relative Read Count Export

Relative eDNA read counts (TSS-normalized) for the target taxon group, restricted to
species in the trawl list, exported as two CSVs:

- **By station** (`_by_station.csv`): one row per `trawl_id × sample_type × species`
- **Trawl-wide** (`_trawl_wide.csv`): mean (± SD) relative read count across stations,
  one row per `sample_type × species`, with counts of how many stations had that
  sample type and how many detected the species

```{r export-relreads}
# Species that are in the trawl lookup for this taxon filter
trawl_species_set <- unique(trawl_agg$species)

output_dir <- if (exists("path_analysis")) path_analysis else "."
base_name  <- paste0("relative_reads_", tolower(gsub(" ", "_", taxon_label)), "_trawl_species")

# ── 1. By-station CSV: one row per trawl_id × sample_type × species ──────────
relreads_by_station <- edna_long %>%
  filter(!is.na(station), !is.na(sample_type)) %>%
  filter(str_to_sentence(str_squish(species)) %in%
           str_to_sentence(str_squish(trawl_species_set))) %>%
  group_by(station, sample_type, species) %>%
  summarise(rel_read_count = sum(reads_tss, na.rm = TRUE), .groups = "drop") %>%
  rename(trawl_id = station) %>%
  arrange(trawl_id, sample_type, species)

cat("By-station export — rows:", nrow(relreads_by_station),
    "| stations:", n_distinct(relreads_by_station$trawl_id),
    "| sample types:", n_distinct(relreads_by_station$sample_type),
    "| species:", n_distinct(relreads_by_station$species), "\n")

fname_station <- paste0(base_name, "_by_station.csv")
write_csv(relreads_by_station, file.path(output_dir, fname_station))
cat("Exported:", fname_station, "\n\n")

# ── 2. Trawl-wide CSV: mean relative read count across stations,
#       one row per sample_type × species ─────────────────────────────────────
# Mean is taken over stations where the species was observed (rel_read_count > 0
# OR the station had that sample_type present, depending on preference).
# Here we average only over stations that had that sample_type, filling
# absent species with 0 so absent detections count against the mean.
all_combos <- relreads_by_station %>%
  distinct(trawl_id, sample_type) %>%
  crossing(species = unique(relreads_by_station$species))

relreads_wide <- all_combos %>%
  left_join(relreads_by_station, by = c("trawl_id", "sample_type", "species")) %>%
  mutate(rel_read_count = replace_na(rel_read_count, 0)) %>%
  group_by(sample_type, species) %>%
  summarise(
    n_stations          = n(),
    n_stations_detected = sum(rel_read_count > 0),
    mean_rel_read_count = mean(rel_read_count),
    sd_rel_read_count   = sd(rel_read_count),
    .groups = "drop"
  ) %>%
  arrange(sample_type, desc(mean_rel_read_count))

cat("Trawl-wide export — rows:", nrow(relreads_wide),
    "| sample types:", n_distinct(relreads_wide$sample_type),
    "| species:", n_distinct(relreads_wide$species), "\n")

fname_wide <- paste0(base_name, "_trawl_wide.csv")
write_csv(relreads_wide, file.path(output_dir, fname_wide))
cat("Exported:", fname_wide, "\n")

# ── Preview tables ────────────────────────────────────────────────────────────
relreads_by_station %>%
  mutate(rel_read_count = round(rel_read_count, 5)) %>%
  head(15) %>%
  kable(caption = paste("By-station relative eDNA read counts —", taxon_label,
                        "(first 15 rows)"),
        col.names = c("Trawl ID", "Sample type", "Species", "Rel. read count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  print()

relreads_wide %>%
  mutate(across(c(mean_rel_read_count, sd_rel_read_count), ~ round(.x, 5))) %>%
  kable(caption = paste("Trawl-wide mean relative eDNA read counts —", taxon_label),
        col.names = c("Sample type", "Species", "N stations",
                      "N detected", "Mean rel. count", "SD")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  print()
```

---

# 9. Session Info

```{r session-info}
sessionInfo()
```
